### 2021-12-02 paper/paper.tex
title and abstract

\documentclass[9pt,twocolumn,twoside]{gsajnl_modified}
% Use the documentclass option 'lineno' to view line numbers

\usepackage[htt]{hyphenat}  % https://tex.stackexchange.com/a/543
\usepackage[export]{adjustbox}
\usepackage{xurl}
\usepackage{stfloats}

\renewcommand{\topfraction}{0.9}	% max fraction of floats at top
    \renewcommand{\bottomfraction}{0.8}	% max fraction of floats at bottom
    %   Parameters for TEXT pages (not float pages):
    \setcounter{topnumber}{2}
    \setcounter{bottomnumber}{2}
    \setcounter{totalnumber}{4}     % 2 may work better
    \setcounter{dbltopnumber}{2}    % for 2-column pages
    \renewcommand{\dbltopfraction}{0.9}	% fit big float above 2-col. text
    \renewcommand{\textfraction}{0.07}	% allow minimal text w. figs
    %   Parameters for FLOAT pages (not text pages):
    \renewcommand{\floatpagefraction}{0.7}	% require fuller float pages
	% N.B.: floatpagefraction MUST be less than topfraction !!
    \renewcommand{\dblfloatpagefraction}{0.7}	% require fuller float pages

\title{An antibody-escape calculator for mutations to the SARS-CoV-2 receptor-binding domain}

\author[*]{\Large Allison J. Greaney$^{1,2,3}$, Tyler N. Starr$^{1,4}$, Jesse D. Bloom$^{1,2,4}$}

\affil[1]{Basic Sciences and Computational Biology, Fred Hutchinson Cancer Center

} 
\affil[2]{Department of Genome Sciences, University of Washington

}
\affil[3]{Medical Scientist Training Program, University of Washington

}
\affil[4]{Howard Hughes Medical Institute

Seattle, WA, USA
}

\keywords{}

\runningtitle{} % For use in the footer 
\runningauthor{}

\begin{abstract}
A key goal of SARS-CoV-2 surveillance is to rapidly identify viral variants with mutations that reduce neutralization by polyclonal antibodies elicited by vaccination or infection.
Unfortunately, direct experimental characterization of new viral variants substantially lags their sequence-based identification. 
Here we help overcome this challenge by providing an ``escape calculator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
Specifically, the calculator aggregates deep mutational scanning data to visualize how mutations impact polyclonal antibody recognition.
It also quantitatively scores how much each mutant is expected to escape RBD-targeted antibody neutralization.
These scores correlate with neutralization assays performed by multiple labs on a range of SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the mutations in the recently described Omicron variant.
An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
\end{abstract}

\begin{document}

\maketitle
\thispagestyle{firststyle}
%\marginmark
\firstpagefootnote

\correspondingauthoraffiliation{}{*Corresponding author: \href{mailto:jbloom@fredhutch.org}{jbloom@fredhutch.org}}
\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page

\lettrine[lines=2]{\color{color2}T}{}he last two years...

\section{Results}

\subsection{}

\section{Discussion}

{\small

\section{Methods}
\subsection{Code and data availability}

\section{Acknowledgments}
{\color{red} More}
JDB is an Investigator of the Howard Hughes Medical Institute.

\section{Competing interests}
JDB consults for Moderna and Flagship Labs 77, and is an inventor on a Fred Hutch licensed patents related to deep mutational scanning of viral proteins.

}

\bibliography{references}

\end{document}

### 2021-12-02 paper/paper.tex
draft of intro

@@
 
-\lettrine[lines=2]{\color{color2}T}{}he last two years...
+\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo rapid antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
+This antigenic evolution is already apparent for SARS-CoV-2, as a variety of new viral variants with reduced antibody neutralization have already emerged only $\sim$2 years since the virus first started to spread in humans~\citep{?}.
+A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{?}.
+Unfortunately, the rate at which new variants arise and spread can outstrip the speed with which these laborious experiments can be performed.
+
+A partial solution is to use deep mutational scanning experiments to \emph{prospectively} characterize how viral mutations impact antibody binding or neutralization.
+These experiments can systematically map the antigenic impacts of all possible amino-acid mutations in key regions of spike on specific antibodies or serum samples~\citep{?}.
+However, SARS-CoV-2 variants of concern typically have many mutations in spike, and it is not feasible to experimentally characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.
+
+Here we provide a partial solution to this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of combinations of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{?}.
+The resulting ``escape calculator'' enables both qualitative visualization and quantitative scoring of the antigenic impacts of arbitrary combinations of RBD mutations.
 
@@
 \section{Acknowledgments}
-{\color{red} More}
+{\color{red} GISAID}
 JDB is an Investigator of the Howard Hughes Medical Institute.

### 2021-12-03 paper/paper.tex
added first results section

--- 2021-12-02 paper/paper.tex
@@
 \usepackage{stfloats}
+\usepackage{sidecap}
 
@@
 A key goal of SARS-CoV-2 surveillance is to rapidly identify viral variants with mutations that reduce neutralization by polyclonal antibodies elicited by vaccination or infection.
-Unfortunately, direct experimental characterization of new viral variants substantially lags their sequence-based identification. 
-Here we help overcome this challenge by providing an ``escape calculator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
+Unfortunately, direct experimental characterization of new viral variants lags their sequence-based identification. 
+Here we help address this challenge by providing an ``escape calculator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
 Specifically, the calculator aggregates deep mutational scanning data to visualize how mutations impact polyclonal antibody recognition.
-It also quantitatively scores how much each mutant is expected to escape RBD-targeted antibody neutralization.
-These scores correlate with neutralization assays performed by multiple labs on a range of SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the mutations in the recently described Omicron variant.
+It also scores how much each mutant is expected to escape RBD-targeted antibody neutralization.
+These scores correlate with neutralization assays performed by multiple labs on a range of SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
 An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
@@
 
-\lettrine[lines=2]{\color{color2}T}{}he last two years...
+\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo rapid antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
+This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have already emerged only $\sim$2 years since the virus first started to spread in humans~\citep{?}.
+A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{?}.
+Unfortunately, the rate at which new variants arise outstrips the speed with which these experiments can be performed.
+
+A partial solution is to use deep mutational scanning to \emph{prospectively} characterize how viral mutations impact antibody binding or neutralization.
+Deep mutational scanning can systematically the antigenic impacts of all possible amino-acid mutations in key regions of spike on specific antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
+However, SARS-CoV-2 variants of concern typically have multiple mutations in spike, and it is not feasible to characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.
+
+Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of combinations of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{?}.
+The resulting ``escape calculator'' enables both qualitative visualization and quantitative scoring of the antigenic impacts of arbitrary combinations of RBD mutations.
+Importantly, the escape calculator is based on simple and intuitive transformations of direct experimental measurements, and does not involve complex black-box computational methods.
 
@@
 
-\subsection{}
+\subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
+A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody.
+This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
+Here we will work with these site-level escape maps.
+
+As a small example to illustrate the principle behind our approach, Figure~\ref{fig:mini_example}A shows previously reported measurements~\citep{starr2021prospective, starr2021complete} of how mutations to each RBD site affect binding by three monoclonal antibodies: LY-CoV016 (etesevimab), LY-CoV555 (bamlanivimab), and REGN10987 (imdevimab).
+Each antibody targets a different one of the three main neutralizing epitopes on the RBD: LY-CoV016 targets the class 1 epitope, LY-CoV555 the class 2 epitope, and REGN10987 the class 3 epitope~\citep{barnes2020sars,greaney2021mapping}.
+Because the antibodies have distinct epitopes, they are each escaped by largely distinct sets of mutations: LY-CoV016 is most strongly escaped by mutations at site K417, LY-CoV555 at site E484, and REGN10987 at sites K444 to G446 (Figure~\ref{fig:mini_example}A).
+
+Now imagine a polyclonal antibody mix consisting of these three antibodies combined at equal potencies.
+We can generate a site-level ``escape map'' for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
+Because this polyclonal escape map is the average of the three monoclonal antibody escape maps, its largest peaks are at the sites of strongest escape for each of the individual antibodies: sites K417, E484, and K444 to G446.
+
+Next imagine removing one antibody from the mix by mutating its epitope.
+Figure~\ref{fig:mini_example}B shows the resulting escape map if LY-CoV555 is ablated, as would occur if site E484 was mutated.
+The thick black line for the antibody mix no longer has peaks at E484 and other sites targeted by LY-CoV555, such as F490.
+Therefore, in this hypothetical polyclonal antibody mix, escape at sites E484 and F490 is correlated since both are targeted by the same antibody.
+However, the polyclonal mix's escape map at sites K417 and N460 is unaffected by mutations that escape LY-CoV555, since they are targeted by a different antibody, LY-CoV016.
+But if we also ablate LY-CoV016 (such as by mutating site K417), then the peaks at K417 and N460 also disappear, and the most remaining peaks are at sites targeted by REGN10987, such as K444 to G446 (Figure~\ref{fig:mini_example}C).
+Of course, if REGN10987 was also ablated such as by mutating site G446, then the polyclonal antibody mix would have no remaining activity.
+(This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.)
+
+\begin{SCfigure*}[0.3]
+\includegraphics[width=0.71\textwidth]{figures/mini_example/mini_example.pdf}
+
+\caption{
+Escape map for a hypothetical polyclonal mix consisting of an equipotent mixture of three monoclonal antibodies targeting distinct epitopes on the SARS-CoV-2 RBD.
+(A) Experimentally measured escape maps for three antibodies, and the mean of these maps (thick black line).
+Each point on the x-axis represents a site in the RBD, and the y-axis represents the total measured escape by all mutations at that site scaled so the maximum for each antibody is one.
+(B) Escape map if the contribution of antibody LY-CoV555 is ablated.
+(C) Escape map if the contributions of antibodies LY-CoV555 and LY-CoV016 are ablated.
+An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.}
+\label{fig:mini_example}
+\end{SCfigure*}
 

### 2021-12-03 paper/paper.tex
second Results section

@@
 \usepackage{stfloats}
-\usepackage{sidecap}
+\usepackage[leftcaption]{sidecap}
+\sidecaptionvpos{figure}{t}
 
@@
 It also scores how much each mutant is expected to escape RBD-targeted antibody neutralization.
-These scores correlate with neutralization assays performed by multiple labs on a range of SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
+These scores correlate with neutralization assays performed by multiple labs on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
 An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
@@
 Each antibody targets a different one of the three main neutralizing epitopes on the RBD: LY-CoV016 targets the class 1 epitope, LY-CoV555 the class 2 epitope, and REGN10987 the class 3 epitope~\citep{barnes2020sars,greaney2021mapping}.
-Because the antibodies have distinct epitopes, they are each escaped by largely distinct sets of mutations: LY-CoV016 is most strongly escaped by mutations at site K417, LY-CoV555 at site E484, and REGN10987 at sites K444 to G446 (Figure~\ref{fig:mini_example}A).
+Because the antibodies have distinct epitopes, they are each escaped by largely distinct sets of mutations: LY-CoV016 is most strongly escaped by mutations at site 417, LY-CoV555 at site 484, and REGN10987 at sites 444 to 446 (Figure~\ref{fig:mini_example}A).
 
@@
 We can generate a site-level ``escape map'' for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
-Because this polyclonal escape map is the average of the three monoclonal antibody escape maps, its largest peaks are at the sites of strongest escape for each of the individual antibodies: sites K417, E484, and K444 to G446.
+Because this polyclonal escape map is the average of the three monoclonal antibody escape maps, its largest peaks are at the sites of strongest escape for each of the individual antibodies: sites 417, 484, and 444 to 446.
 
 Next imagine removing one antibody from the mix by mutating its epitope.
-Figure~\ref{fig:mini_example}B shows the resulting escape map if LY-CoV555 is ablated, as would occur if site E484 was mutated.
-The thick black line for the antibody mix no longer has peaks at E484 and other sites targeted by LY-CoV555, such as F490.
+Figure~\ref{fig:mini_example}B shows the resulting escape map if LY-CoV555 is ablated, as would occur if site 484 was mutated.
+The thick black line for the antibody mix no longer has peaks at E484 and other sites targeted by LY-CoV555, such as 490.
 Therefore, in this hypothetical polyclonal antibody mix, escape at sites E484 and F490 is correlated since both are targeted by the same antibody.
 However, the polyclonal mix's escape map at sites K417 and N460 is unaffected by mutations that escape LY-CoV555, since they are targeted by a different antibody, LY-CoV016.
-But if we also ablate LY-CoV016 (such as by mutating site K417), then the peaks at K417 and N460 also disappear, and the most remaining peaks are at sites targeted by REGN10987, such as K444 to G446 (Figure~\ref{fig:mini_example}C).
-Of course, if REGN10987 was also ablated such as by mutating site G446, then the polyclonal antibody mix would have no remaining activity.
+But if we also ablate LY-CoV016 (such as by mutating site 417), then the peaks at 417 and 460 also disappear, and the most remaining peaks are at sites targeted by REGN10987, such as 444 to 446 (Figure~\ref{fig:mini_example}C).
+Of course, if REGN10987 was also ablated such as by mutating site 446, then the polyclonal antibody mix would have no remaining activity.
 (This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.)
 
-\begin{SCfigure*}[0.3]
+\begin{SCfigure*}[][]
 \includegraphics[width=0.71\textwidth]{figures/mini_example/mini_example.pdf}
@@
 Escape map for a hypothetical polyclonal mix consisting of an equipotent mixture of three monoclonal antibodies targeting distinct epitopes on the SARS-CoV-2 RBD.
-(A) Experimentally measured escape maps for three antibodies, and the mean of these maps (thick black line).
+{\bf (A)} Experimentally measured escape maps for three antibodies, and the mean of these maps (thick black line).
 Each point on the x-axis represents a site in the RBD, and the y-axis represents the total measured escape by all mutations at that site scaled so the maximum for each antibody is one.
-(B) Escape map if the contribution of antibody LY-CoV555 is ablated.
-(C) Escape map if the contributions of antibodies LY-CoV555 and LY-CoV016 are ablated.
+{\bf (B)} Escape map if the contribution of antibody LY-CoV555 is ablated.
+{\bf (C)} Escape map if the contributions of antibodies LY-CoV555 and LY-CoV016 are ablated.
 An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.}
@@
 \end{SCfigure*}
+
+\subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape calculator}
+%The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
+To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing monoclonal antibodies elicited by SARS-CoV-2.
+These antibodies were isolated from a variety of patient cohorts within the first year of the pandemic (see Methods for details).
+An assumption of the analysis that follows is that an equipotent mixture of these 33 antibodies represents the neutralizing activity of human sera; we emphasize that this assumption is imperfect since in reality the antibodies were chosen for prior study for a variety of ad hoc reasons.
+The escape maps for all the individual antibodies can be interactively interrogated at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/}.
+
+\begin{SCfigure*}[][]
+\includegraphics[width=0.8\textwidth]{figures/escape_calc/escape_calc.pdf}
+\caption{
+An escape calculator generated by aggregating deep mutational scanning for 33 neutralizing antibodies targeting the SARS-CoV-2 RBD.
+{\bf (A)} The blue line shows the extent of escape mediated by mutations at each site, as estimated by simply averaging the data for all the individual antibodies.
+{\bf (B)} The blue line shows escape map after a mutation to site 484 (red point) ablates recognition by antibodies strongly targeting that site, while the gray line shows the original escape map in the absence of any mutations.
+{\bf (C)} The escape map after mutating sites 417, 484, and 501 (the three RBD sites mutated in the Beta variant).
+An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}.}
+\label{fig:escape_calc}
+\end{SCfigure*}
+
+The overall polyclonal escape map generated by averaging the data for all 33 antibodies is in Figure \ref{fig:escape_calc}A.
+As in the toy three-antibody example in the previous subsection, there are peaks at  sites 417, 444 to 446, and 484.
+However, the peak at E484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are immunodominant in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping}.
+In addition, there are smaller peaks at a variety of sites, reflecting the fact that each antibody has a somewhat idiosyncratic epitope that is slightly differently affected by mutations at specific sites (Figure \ref{fig:escape_calc}A).
+
+We can follow the principle outlined in the toy example in the previous subsection to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
+Specifically, we reduce the contribution of each antibody by an amount that scales with how strongly that antibody targets each mutated site (see Methods for mathematical details).
+For instance, the blue lines in Figure~\ref{fig:escape_calc}B show the polyclonal escape map after mutating site 484.
+Mutating site 484 obviously drops the contribution of that site, but it also decreases the contribution of other sites such as 490 that are commonly targeted by antibodies with epitopes that include site 484.
+In contrast, mutating site 484 has minimal effect on the polyclonal escape map at sites like 417 or 444 to 446, since those sites are generally targeted by antibodies that are unaffected by mutations at site 484.
+
+We can also calculate the expected effects of compound mutations.
+Figure~\ref{fig:escape_calc}C shows the polyclonal escape map after mutating all three RBD sites that are changed in the Beta variant (sites 417, 484, and 501).
+This polyclonal escape map has lost contributions not only from the mutated sites, but also sites that tend to form common epitopes with either 417 or 484 (e.g., sites 455, 456, 486, and 490).
+However, the escape map still has major contributions from antibodies targeting sites like 444 to 446, since such antibodies are generally unaffected by mutations at sites 417, 484, or 501.
+
+We recommend the reader explores the interactive escape calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, to perform calculations like those in Figure~\ref{fig:escape_calc} for arbitrary combinations of mutated RBD sites.
+Such visual exploration of different combinations of mutations provides an intuitive sense of the antigenic structure of the RBD.
+
+\subsection{}
+Omicron is 0.2.
+PMS20 is 0.19
 
@@
 \subsection{Code and data availability}
+
+\subsection{Mutations in Omicron variant and spike polymutant PMS20}
+Omicron has mutations at the following 15 RBD sites: 339, 371, 373, 375, 417, 440, 446, 477, 478, 484, 493, 496, 498, 501, and 505.
+PMS20 has mutations at the following 8 RBD sites: 346, 417, 440, 445, 455, 475, 484, and 501.
 

### 2021-12-03 paper/paper.tex
third results subsection

@@
 These scores correlate with neutralization assays performed by multiple labs on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
-An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
+An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a module for batch processing.
 \end{abstract}
@@
 
-\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo rapid antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
-This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have already emerged only $\sim$2 years since the virus first started to spread in humans~\citep{?}.
-A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{?}.
+\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
+This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have emerged only $\sim$2 years since the virus first started to spread in humans~\citep{?}.
+A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
 Unfortunately, the rate at which new variants arise outstrips the speed with which these experiments can be performed.
@@
 A partial solution is to use deep mutational scanning to \emph{prospectively} characterize how viral mutations impact antibody binding or neutralization.
-Deep mutational scanning can systematically the antigenic impacts of all possible amino-acid mutations in key regions of spike on specific antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
+Deep mutational scanning can systematically measure the antigenic impacts of all possible amino-acid mutations in key regions of spike on specific antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
 However, SARS-CoV-2 variants of concern typically have multiple mutations in spike, and it is not feasible to characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.
 
-Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of combinations of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{?}.
-The resulting ``escape calculator'' enables both qualitative visualization and quantitative scoring of the antigenic impacts of arbitrary combinations of RBD mutations.
-Importantly, the escape calculator is based on simple and intuitive transformations of direct experimental measurements, and does not involve complex black-box computational methods.
+Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of combinations of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
+The resulting ``escape calculator'' enables qualitative visualization and quantitative scoring of the antigenic impacts of arbitrary combinations of mutations.
+Importantly, the escape calculator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.
 
@@
 \subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
-A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody.
+A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}.
 This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
@@
 As a small example to illustrate the principle behind our approach, Figure~\ref{fig:mini_example}A shows previously reported measurements~\citep{starr2021prospective, starr2021complete} of how mutations to each RBD site affect binding by three monoclonal antibodies: LY-CoV016 (etesevimab), LY-CoV555 (bamlanivimab), and REGN10987 (imdevimab).
-Each antibody targets a different one of the three main neutralizing epitopes on the RBD: LY-CoV016 targets the class 1 epitope, LY-CoV555 the class 2 epitope, and REGN10987 the class 3 epitope~\citep{barnes2020sars,greaney2021mapping}.
-Because the antibodies have distinct epitopes, they are each escaped by largely distinct sets of mutations: LY-CoV016 is most strongly escaped by mutations at site 417, LY-CoV555 at site 484, and REGN10987 at sites 444 to 446 (Figure~\ref{fig:mini_example}A).
+Each antibody targets a different epitope on the RBD: LY-CoV016 targets the class 1 epitope, LY-CoV555 the class 2 epitope, and REGN10987 the class 3 epitope~\citep{barnes2020sars,greaney2021mapping}.
+Because the antibodies have distinct epitopes, they are escaped by largely distinct sets of mutations: LY-CoV016 is most strongly escaped by mutations at site 417, LY-CoV555 at site 484, and REGN10987 at sites 444--446 (Figure~\ref{fig:mini_example}A).
 
@@
 We can generate a site-level ``escape map'' for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
-Because this polyclonal escape map is the average of the three monoclonal antibody escape maps, its largest peaks are at the sites of strongest escape for each of the individual antibodies: sites 417, 484, and 444 to 446.
+Because this polyclonal escape map is the average of the three monoclonal antibody escape maps, its largest peaks are at the sites of strongest escape for each individual antibody: sites 417, 484, and 444--446.
 
@@
 Figure~\ref{fig:mini_example}B shows the resulting escape map if LY-CoV555 is ablated, as would occur if site 484 was mutated.
-The thick black line for the antibody mix no longer has peaks at E484 and other sites targeted by LY-CoV555, such as 490.
-Therefore, in this hypothetical polyclonal antibody mix, escape at sites E484 and F490 is correlated since both are targeted by the same antibody.
-However, the polyclonal mix's escape map at sites K417 and N460 is unaffected by mutations that escape LY-CoV555, since they are targeted by a different antibody, LY-CoV016.
-But if we also ablate LY-CoV016 (such as by mutating site 417), then the peaks at 417 and 460 also disappear, and the most remaining peaks are at sites targeted by REGN10987, such as 444 to 446 (Figure~\ref{fig:mini_example}C).
+The thick black line for the antibody mix no longer has peaks at 484 and other sites targeted by LY-CoV555, such as 490.
+Therefore, in this hypothetical polyclonal antibody mix, escape at sites 484 and 490 is correlated since both are targeted by the same antibody.
+However, the polyclonal mix's escape map at sites 417 and 460 is unaffected by mutations that escape LY-CoV555, since they are targeted by a different antibody, LY-CoV016.
+But if we also ablate LY-CoV016 (such as by mutating site 417), then the peaks at 417 and 460 also disappear, and the remaining peaks are at sites targeted by REGN10987, such as 444--446 (Figure~\ref{fig:mini_example}C).
 Of course, if REGN10987 was also ablated such as by mutating site 446, then the polyclonal antibody mix would have no remaining activity.
-(This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.)
+This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.
 
@@
 %The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
-To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing monoclonal antibodies elicited by SARS-CoV-2.
+To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing antibodies elicited by SARS-CoV-2.
 These antibodies were isolated from a variety of patient cohorts within the first year of the pandemic (see Methods for details).
@@
 The overall polyclonal escape map generated by averaging the data for all 33 antibodies is in Figure \ref{fig:escape_calc}A.
-As in the toy three-antibody example in the previous subsection, there are peaks at  sites 417, 444 to 446, and 484.
-However, the peak at E484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are immunodominant in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping}.
-In addition, there are smaller peaks at a variety of sites, reflecting the fact that each antibody has a somewhat idiosyncratic epitope that is slightly differently affected by mutations at specific sites (Figure \ref{fig:escape_calc}A).
+As in the toy three-antibody example in the previous subsection, there are peaks at  sites 417, 484, and 444--446.
+However, the peak at 484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are immunodominant in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping}.
+In addition, there are smaller peaks at a variety of other sites, reflecting the fact that each antibody has a somewhat idiosyncratic epitope (Figure \ref{fig:escape_calc}A).
 
@@
 Mutating site 484 obviously drops the contribution of that site, but it also decreases the contribution of other sites such as 490 that are commonly targeted by antibodies with epitopes that include site 484.
-In contrast, mutating site 484 has minimal effect on the polyclonal escape map at sites like 417 or 444 to 446, since those sites are generally targeted by antibodies that are unaffected by mutations at site 484.
+In contrast, mutating site 484 has minimal effect on the polyclonal escape map at sites like 417 or 444--446, since those sites are generally targeted by antibodies that are unaffected by mutations at site 484.
 
@@
 Figure~\ref{fig:escape_calc}C shows the polyclonal escape map after mutating all three RBD sites that are changed in the Beta variant (sites 417, 484, and 501).
-This polyclonal escape map has lost contributions not only from the mutated sites, but also sites that tend to form common epitopes with either 417 or 484 (e.g., sites 455, 456, 486, and 490).
-However, the escape map still has major contributions from antibodies targeting sites like 444 to 446, since such antibodies are generally unaffected by mutations at sites 417, 484, or 501.
+This polyclonal escape map has lost contributions not only from the mutated sites, but also sites that form common epitopes with 417 or 484 (e.g., sites 455, 456, 486, and 490).
+However, the escape map still has major contributions from antibodies targeting sites like 444--446, since such antibodies are generally unaffected by mutations at sites 417, 484, or 501.
 
@@
 Such visual exploration of different combinations of mutations provides an intuitive sense of the antigenic structure of the RBD.
+
+\subsection{The escape calculations correlate well with neutralization assays on SARS-CoV-2 variants}
+We can also define a quantitative score that summarizes the polyclonal antibody activity that remains after an arbitrary combination of RBD mutations.
+This score is defined using the same principle as the site-wise escape calculator in the previous subsection: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain after the mutations (see Methods for details).
+This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).
+
+\begin{SCfigure*}[][]
+\includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
+\caption{Correlation of antibody binding scores from the escape calculator with experimentally measured fold changes in neutralization IC50 values for SARS-CoV-2 variants and mutants.
+The experimental data of \citet{lucas2021impact} was generated using live virus and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
+The experimental data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with the vaccine sera from the Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively.
+The plotted fold changes in neutralization represent the geometric means over all subjects in each cohort.
+}
+\label{fig:neut_studies}
+\end{SCfigure*}
+
+To test how these escape-calculator scores compare to actual experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
+One can imagine many reasons why the escape-calculator scores might differ from the real neutralization titers: the calculator only considers RBD mutations and neglects those in other domains such as the NTD, the antibodies used by the calculator might not accurately reflect the real mix in polyclonal sera, etc.
+But despite all these potential caveats, the escape-calculator scores correlate quite well with the measured neutralization titers across all studies and cohorts (Figure~\ref{fig:neut_studies}).
+Therefore, the simple and intuitive approach used by the calculator seems to accurately reflect the dominant features of polyclonal antibody escape in the RBD.
 

### 2021-12-04 paper/paper.tex
plot variants

@@
 \includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
-\caption{Correlation of antibody binding scores from the escape calculator with experimentally measured fold changes in neutralization IC50 values for SARS-CoV-2 variants and mutants.
+\caption{Correlation of escape scores from the calculator with experimentally measured fold changes in neutralization IC50 values for SARS-CoV-2 variants and mutants.
 The experimental data of \citet{lucas2021impact} was generated using live virus and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.

### 2021-12-04 paper/paper.tex
start making Omicron figures
added Omicron results subsection

@@
 \subsection{The escape calculations correlate well with neutralization assays on SARS-CoV-2 variants}
-We can also define a quantitative score that summarizes the polyclonal antibody activity that remains after an arbitrary combination of RBD mutations.
+We can also define a quantitative score that summarizes the polyclonal antibody binding that remains after an arbitrary combination of RBD mutations.
 This score is defined using the same principle as the site-wise escape calculator in the previous subsection: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain after the mutations (see Methods for details).
@@
 
-\subsection{}
-Omicron is 0.2.
-PMS20 is 0.19
+\subsection{The escape calculator suggests extensive antigenic change in the new Omicron variant}
+We applied the escape calculator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{?}.
+The overall escape-calculator binding score for the polyclonal antibody mix is much lower for the Omicron variant than any other SARS-CoV-2 variants of concern (Figure~\ref{fig:Omicron}A).
+The Omicron variant's calculated score is roughly equivalent to that of a polymutant spike (PMS20) that was artificially engineered in a pseudovirus by \citet{schmidt2021high} to maximize escape from polyclonal serum antibodies.
+For comparison, \citet{schmidt2021high} measured that neutralization titers against this artificial PMS20 spike were reduced by $\sim$20- to $\sim$80-fold for sera from various cohorts of vaccinated and infected individuals.
+
+\begin{figure*}
+\includegraphics[width=\linewidth]{figures/Omicron/Omicron.pdf}
+\caption{Escape calculations for the Omicron variant.
+{\bf (A)} The calculated binding scores for SARS-CoV-2 variants and the artificial polymutant spike (PMS20) generated by \citet{schmidt2021high}.
+Scores of one indicate no mutations affect binding, and scores of zero indicate no antibody binding remains.
+{\bf (B)} The calculated escape map for the Omicron variant's RBD (blue) compared to an unmutated RBD (gray), with sites of mutations in the Omicron variant in red.
+}
+\label{fig:Omicron}
+\end{figure*}
+
+The site-level escape map for the Omicron variant's RBD is shown in Figure~\ref{fig:Omicron}B.
+That map shows that the Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
+Exploration of the mutations using the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that the mutations at sites 417, 446, and 484 are the biggest drivers of this antigenic change, although other mutations also contribute.
+The residual peaks in the map suggests that the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.
 

### 2021-12-04 paper/paper.tex
full draft of Discussion

@@
 Unfortunately, direct experimental characterization of new viral variants lags their sequence-based identification. 
-Here we help address this challenge by providing an ``escape calculator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
-Specifically, the calculator aggregates deep mutational scanning data to visualize how mutations impact polyclonal antibody recognition.
-It also scores how much each mutant is expected to escape RBD-targeted antibody neutralization.
-These scores correlate with neutralization assays performed by multiple labs on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
+Here we help address this challenge by aggregating deep mutational scanning data into an ``escape calculator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
+The calculator can be used to intuitively visualize how mutations impact polyclonal antibody recognition, and score the expected antigenic effect of combinations of mutations.
+These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
 An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a module for batch processing.
@@
 \lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
-This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have emerged only $\sim$2 years since the virus first started to spread in humans~\citep{?}.
+This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have emerged only $\sim$2 years since the virus first started to spread in humans.
 A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
@@
 
-A partial solution is to use deep mutational scanning to \emph{prospectively} characterize how viral mutations impact antibody binding or neutralization.
-Deep mutational scanning can systematically measure the antigenic impacts of all possible amino-acid mutations in key regions of spike on specific antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
-However, SARS-CoV-2 variants of concern typically have multiple mutations in spike, and it is not feasible to characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.
+A partial solution is to use deep mutational scanning experiments to \emph{prospectively} measure how viral mutations impact antibody binding or neutralization.
+Deep mutational scanning can systematically measure the antigenic impacts of all possible amino-acid mutations in key regions of spike on monoclonal antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
+However, SARS-CoV-2 variants of concern typically have multiple mutations, and it is not feasible to experimentally characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.
 
-Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of combinations of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
-The resulting ``escape calculator'' enables qualitative visualization and quantitative scoring of the antigenic impacts of arbitrary combinations of mutations.
+Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
+The resulting ``escape calculator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
 Importantly, the escape calculator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.
@@
 
-Now imagine a polyclonal antibody mix consisting of these three antibodies combined at equal potencies.
-We can generate a site-level ``escape map'' for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
-Because this polyclonal escape map is the average of the three monoclonal antibody escape maps, its largest peaks are at the sites of strongest escape for each individual antibody: sites 417, 484, and 444--446.
+Now imagine a polyclonal antibody mix of these three antibodies combined at equal potencies.
+We can generate an escape map for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
+Because this polyclonal escape map is the average of the monoclonal antibody maps, its largest peaks are at the sites of strongest escape for each individual antibody: 417, 484, and 444--446.
 
@@
 The thick black line for the antibody mix no longer has peaks at 484 and other sites targeted by LY-CoV555, such as 490.
-Therefore, in this hypothetical polyclonal antibody mix, escape at sites 484 and 490 is correlated since both are targeted by the same antibody.
+Therefore, in this hypothetical polyclonal antibody mix, escape at sites 484 and 490 is correlated since both sites are targeted by the same antibody.
 However, the polyclonal mix's escape map at sites 417 and 460 is unaffected by mutations that escape LY-CoV555, since they are targeted by a different antibody, LY-CoV016.
@@
 \subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape calculator}
-%The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
+The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
 To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing antibodies elicited by SARS-CoV-2.
@@
 
-The overall polyclonal escape map generated by averaging the data for all 33 antibodies is in Figure \ref{fig:escape_calc}A.
-As in the toy three-antibody example in the previous subsection, there are peaks at  sites 417, 484, and 444--446.
-However, the peak at 484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are immunodominant in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping}.
+The overall polyclonal escape map generated by averaging the experimental data for all 33 antibodies is in Figure \ref{fig:escape_calc}A.
+As in the toy three-antibody example in the previous section, there are peaks at  sites 417, 484, and 444--446.
+However, the peak at 484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are especially common in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
 In addition, there are smaller peaks at a variety of other sites, reflecting the fact that each antibody has a somewhat idiosyncratic epitope (Figure \ref{fig:escape_calc}A).
 
-We can follow the principle outlined in the toy example in the previous subsection to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
-Specifically, we reduce the contribution of each antibody by an amount that scales with how strongly that antibody targets each mutated site (see Methods for mathematical details).
+We can follow the principle outlined in the toy example in the previous section to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
+Specifically, we reduce the contribution of each antibody by an amount that scales with how strongly that antibody targets each mutated site (see Methods for details).
 For instance, the blue lines in Figure~\ref{fig:escape_calc}B show the polyclonal escape map after mutating site 484.
@@
 \subsection{The escape calculations correlate well with neutralization assays on SARS-CoV-2 variants}
-We can also define a quantitative score that summarizes the polyclonal antibody binding that remains after an arbitrary combination of RBD mutations.
-This score is defined using the same principle as the site-wise escape calculator in the previous subsection: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain after the mutations (see Methods for details).
+For each set of RBD mutations, we can define a quantitative score that represents the polyclonal antibody binding that remains after making these mutations.
+This score is defined using the same principle as the site-wise escape calculator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
 This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).
@@
 
-To test how these escape-calculator scores compare to actual experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
-One can imagine many reasons why the escape-calculator scores might differ from the real neutralization titers: the calculator only considers RBD mutations and neglects those in other domains such as the NTD, the antibodies used by the calculator might not accurately reflect the real mix in polyclonal sera, etc.
+To test how these escape-calculator scores compare to experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody,}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
+One can imagine many reasons why the escape-calculator scores might differ from the real neutralization titers: the calculator only considers RBD mutations, the antibodies used by the calculator might not accurately reflect the real mix in polyclonal sera, etc.
 But despite all these potential caveats, the escape-calculator scores correlate quite well with the measured neutralization titers across all studies and cohorts (Figure~\ref{fig:neut_studies}).
@@
 \subsection{The escape calculator suggests extensive antigenic change in the new Omicron variant}
-We applied the escape calculator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{?}.
-The overall escape-calculator binding score for the polyclonal antibody mix is much lower for the Omicron variant than any other SARS-CoV-2 variants of concern (Figure~\ref{fig:Omicron}A).
+We applied the escape calculator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{ngs-sa2021update,deoliveira2021tweet}.
+The calculated binding score for the Omicron variant is much lower than any other SARS-CoV-2 variants of concern, indicating extensive antibody escape (Figure~\ref{fig:Omicron}A).
 The Omicron variant's calculated score is roughly equivalent to that of a polymutant spike (PMS20) that was artificially engineered in a pseudovirus by \citet{schmidt2021high} to maximize escape from polyclonal serum antibodies.
@@
 The site-level escape map for the Omicron variant's RBD is shown in Figure~\ref{fig:Omicron}B.
-That map shows that the Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
-Exploration of the mutations using the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that the mutations at sites 417, 446, and 484 are the biggest drivers of this antigenic change, although other mutations also contribute.
-The residual peaks in the map suggests that the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.
+The Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
+Exploration of the mutations using the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that mutations at sites 484, 446, and 417 are the biggest drivers of this antigenic change, although other mutations also contribute.
+The residual peaks in the map suggests the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.
 
 \section{Discussion}
+We have described an escape calculator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
+The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus synergistic effects on antigenic escape.
+For instance, sites 417, 484, and 490 are all peaks of antibody escape.
+But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have synergistic effects since they generally escape different antibodies.
+
+Another key aspect of our approach is the interactive visual implementation of the escape calculator, so the user can interrogate the effects of mutations (or combinations of mutations) simply by clicking (or shift-clicking) on sites.
+This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the calculator is performing simple transformations directly on experimental data.
+We encourage interactive use of the calculator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
+However, we note that in addition to the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/bindingcalculator.py}.
+
+There are caveats that should be kept in mind when using the escape calculator.
+First, it only assesses the antigenic effects of mutations, and not how they impact other properties such as viral transmissibility.
+Second, the calculator only considers sites in the RBD, and ignores mutations to other regions of spike.
+Third, the calculator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
+Fourth, the calculator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
+Finally, and in our minds most significantly, the calculator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).
+
+Despite all these caveats, the escape calculator yields binding scores that correlate with experimentally measured neutralization titers.
+In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape calculator: variants of concern generally have combinations of mutations calculated to have synergistic effects on antigenic escape (eg, 417 and 484) rather than combinations calculated to have redundant effects (eg, 484 and 490).
+We suspect the calculator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
+
+However, the situation will become more complex over time.
+Currently, most humans with antibodies to SARS-CoV-2 have been exposed to a RBD antigen that is identical or very similar to that of the early Wuhan-Hu-1 strain.
+Therefore, antigenic studies can reasonably define mutations in reference to that RBD, since it is what the antibodies target.
+But as humans are exposed to more diverged RBD variants, it will become difficult to determine what reference to use to define antigenic mutations, since different individuals will have antibodies targeting different RBDs.
+Additionally, differing exposure histories can leave individuals with different antibody specificities~\citep{cobey2017immune}, a process that is already starting to occur for SARS-CoV-2~\citep{greaney2021sars}.
+So in the future, it will be necessary to stratify the data used by the escape calculator by which RBD variant elicited the antibodies, and aggregate data for antibodies that reflect the sera in question.
+For this reason, we expect to continue adding to the data used by the escape calculator, and emphasize that it will change over time from the version described here, although we provide stable links to the current version in the Methods below.
 
@@
 
+\url{https://altair-viz.github.io/} Altair~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.
+
 \subsection{Mutations in Omicron variant and spike polymutant PMS20}
@@
 \section{Acknowledgments}
-{\color{red} GISAID}
+We thank our collaborators in the Crowe lab, Nussenzweig lab, Bjorkman lab, and at Vir Biotechnology for sharing the antibodies that were used in prior published studies to generate the deep mutational scanning data re-analyzed here.
+We thank researchers in South Africa for rapidly sharing sequence information about the Omicron variant to enable the analyses in the last section of the results.
+This work was supported in part by the NIH/NIAID under R01AI141707 (to JDB) and T32AI083203 (to AJG).
+The work was also supported by the Gates Foundation grant INV-004949 to JDB.
+TNS is a Howard Hughes Medical Institute Fellow of the Damon Runyon Cancer Research Foundation.
 JDB is an Investigator of the Howard Hughes Medical Institute.

### 2021-12-04 paper/paper.tex
complete draft of methods

@@
 
-To test how these escape-calculator scores compare to experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody,}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
+To test how these escape-calculator scores compare to experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
 One can imagine many reasons why the escape-calculator scores might differ from the real neutralization titers: the calculator only considers RBD mutations, the antibodies used by the calculator might not accurately reflect the real mix in polyclonal sera, etc.
@@
 \subsection{Code and data availability}
+The most up-to-date code and data used to implement the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps}, and the version described in this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/tree/bioRxiv_v1}.
 
-\url{https://altair-viz.github.io/} Altair~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.
+The data used by the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv}, and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv}.
 
-\subsection{Mutations in Omicron variant and spike polymutant PMS20}
-Omicron has mutations at the following 15 RBD sites: 339, 371, 373, 375, 417, 440, 446, 477, 478, 484, 493, 496, 498, 501, and 505.
-PMS20 has mutations at the following 8 RBD sites: 346, 417, 440, 445, 455, 475, 484, and 501.
+\subsection{Deep mutational scanning data used by the calculator}
+The experimental data used by the escape calculator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
+In total, these studies contain data for 36 monoclonal antibodies.
+Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the calculator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
+The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and where initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on sequences reported by Regeneron, Eli Lilly, or their collaborators~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
+The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).
+
+The deep mutational scanning measures an escape fraction for each tolerated RBD mutation against each antibody, which represents an estimate of how completely that mutation escapes antibody binding~\citep{greaney2021complete}.
+We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions at each site.
+The results reported in this paper use the sums as the site-level metric, although the calculator has an option (\texttt{escape\_metric}) to use the mean instead.
+We also normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
+The results reported in this paper use the normalized site-level metrics, although the calculator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.
+
+\subsection{Calculation of the impact of mutations}
+The escape calculator determines the impact of mutating sites by calculating how much each antibody is escaped by mutations at each site, and adjusting its contribution to the overall polyclonal mix accordingly.
+
+Specifically, for each antibody $a$ we have a deep mutational scanning measurement $x_{a,r}$ of how much mutating $r$ escapes that antibody.
+In the absence of any mutations, the overall escape map shown for the polyclonal mix is simply the mean over all antibodies, $\frac{1}{A}\sum_a x_{a,r}$ where $A$ is the number of antibodies.
+
+Let $\mathcal{M}$ be the set of sites that are mutated.
+Then for each antibody we compute the binding retained as $b_a\left(\mathcal{M}\right) = \left(\prod\limits_{r \in \mathcal{M}} \frac{\max_{r'}\left(x_{a,r'}\right) - x_{a,r}}{\max_{r'}\left(x_{a,r'}\right)}\right)^s$.
+Essentially, this equation means that if the RBD is mutated at a strong site of escape for an antibody $a$, much if the binding is lost (if it is the strongest site of escape, all binding is lost).
+The variable $s$ represents how dramatically binding is lost for mutations at sites of escape that are not the strongest: larger values of $s$ means mutations at even moderate sites of escape reduce binding a lot.
+In this paper we report calculations with $s = 2$, although the calculator has an option (\texttt{mutation\_escape\_strength}) to choose other values.
+We then define the escape map after the mutations $\mathcal{M}$ as $\frac{1}{A}\sum_a x_{a,r} b_a\left(\mathcal{M}\right)$.
+The calculator shows the escape map with no mutations in gray, and that after mutations in blue.
+
+The overall antibody binding scores represent the fraction of antibodies that still bind, and are calculated simply as $\frac{1}{A}\sum_a b_a\left(\mathcal{M}\right)$.
+
+\subsection{Implementation of interactive calculator}
+The interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} is imlemented using Altair (\url{https://altair-viz.github.io/})~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.
+
+\subsection{Python module with batch-mode calculator}
+A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/bindingcalculator.py}, and has all the same options as the interactive calculator.
+
+\subsection{Compilation of neutralization titers from the literature}
+For Figure~\ref{fig:neut_studies}, we compiled neutralization data from three published studies on SARS-CoV-2 variants and mutants~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}.
+For each study cohort, we computed the geometric mean fold change in neutralization titer over all subjects.
+The numerical compiled data are at \url{https://github.com/jbloomlab/RBD_escape_calculator_paper/tree/main/results/neut_studies}.
+
+\subsection{Mutations in SARS-CoV-2 variants}
+For Figure~\ref{fig:Omicron}, the definitions of which RBD sites are mutated in each variant are shown in Table~\ref{tab:variants}.
+\begin{table}[h]
+{\footnotesize
+\begin{tabular}{ll}
+\toprule
+   variant &                                           mutated RBD sites \\
+\midrule
+Wuhan-Hu-1 &                                                             \\
+     Alpha &                                                         501 \\
+      Beta &                                                 417 484 501 \\
+     Gamma &                                                 417 484 501 \\
+     Delta &                                                     452 478 \\
+    Lambda &                                                     452 490 \\
+        Mu &                                                 346 484 501 \\
+   Omicron & 339 371 373 375 417 440 446 477 478 484 493 496 498 501 505 \\
+     PMS20 &                             346 417 440 445 455 475 484 501 \\
+\bottomrule
+\end{tabular}
+}
+\caption{Mutated RBD sites in SARS-CoV-2 variants.}
+\label{tab:variants}
+\end{table}
 
 \section{Acknowledgments}
-We thank our collaborators in the Crowe lab, Nussenzweig lab, Bjorkman lab, and at Vir Biotechnology for sharing the antibodies that were used in prior published studies to generate the deep mutational scanning data re-analyzed here.
-We thank researchers in South Africa for rapidly sharing sequence information about the Omicron variant to enable the analyses in the last section of the results.
+We thank collaborators in the Crowe lab, Nussenzweig lab, Bjorkman lab, and at Vir Biotechnology for sharing the antibodies that were used in prior published studies to generate the deep mutational scanning data aggregated here.
+We thank the Moyo and de Oliveira labs and other researchers in South Africa and Botswana for rapidly sharing information about the Omicron variant to enable the analyses in the last section of the results.
 This work was supported in part by the NIH/NIAID under R01AI141707 (to JDB) and T32AI083203 (to AJG).
@@
 \section{Competing interests}
-JDB consults for Moderna and Flagship Labs 77, and is an inventor on a Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
-
-}
+JDB consults for Moderna, Flagship Labs 77 and Oncorus.
+JDB is an inventor on a Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
 
 \bibliography{references}
-
+}
 \end{document}

### 2021-12-04 paper/paper.tex
full draft of paper

@@
 These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
-An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a module for batch processing.
+An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
 \end{abstract}
@@
 {\bf (B)} The calculated escape map for the Omicron variant's RBD (blue) compared to an unmutated RBD (gray), with sites of mutations in the Omicron variant in red.
+The mutated RBD sites for each variant are in Table~\ref{tab:variants}.
 }
@@
 The data used by the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv}, and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv}.
+
+A downloadable HTML version of the escape calculator is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/raw/main/docs/_includes/escape_calc_chart.html}, with the version described in this paper at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/docs/_includes/escape_calc_chart.html}.
 

### 2021-12-04 paper/paper.tex
add Tyler edits

@@
 
-\subsection{The escape calculations correlate well with neutralization assays on SARS-CoV-2 variants}
+\subsection{The escape calculations correlate well with neutralization assays of human polyclonal sera against SARS-CoV-2 variants}
 For each set of RBD mutations, we can define a quantitative score that represents the polyclonal antibody binding that remains after making these mutations.
@@
 \includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
-\caption{Correlation of escape scores from the calculator with experimentally measured fold changes in neutralization IC50 values for SARS-CoV-2 variants and mutants.
-The experimental data of \citet{lucas2021impact} was generated using live virus and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
-The experimental data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with the vaccine sera from the Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively.
-The plotted fold changes in neutralization represent the geometric means over all subjects in each cohort.
+\caption{Correlation of calculated binding scores with experimentally measured fold changes in neutralization for SARS-CoV-2 variants and mutants.
+The data of \citet{lucas2021impact} was generated using live virus and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
+The data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with vaccine sera from Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively.
+The fold changes represent the geometric means over all subjects in each cohort.
+An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html}.
 }
@@
 Scores of one indicate no mutations affect binding, and scores of zero indicate no antibody binding remains.
+An interactive version of this plot that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/variants.html}.
 {\bf (B)} The calculated escape map for the Omicron variant's RBD (blue) compared to an unmutated RBD (gray), with sites of mutations in the Omicron variant in red.
@@
 We encourage interactive use of the calculator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
-However, we note that in addition to the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/bindingcalculator.py}.
+However, we note that in addition to the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.
 
@@
 
+\subsection{Interactive versions of figures}
+Interactive versions of all of the figures in this paper are at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/}.
+These figures allow mousing over points to see details, etc.
+
 \subsection{Deep mutational scanning data used by the calculator}
@@
 Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the calculator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
-The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and where initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on sequences reported by Regeneron, Eli Lilly, or their collaborators~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
+The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on sequences reported by Regeneron, Eli Lilly, or their collaborators~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
 The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).
@@
 \subsection{Python module with batch-mode calculator}
-A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/bindingcalculator.py}, and has all the same options as the interactive calculator.
+A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}, and has all the same options as the interactive calculator.
 

### 2022-03-06 paper/reviewer_response.tex
set up to address reviewer comments

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{color}
\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex

\usepackage{amssymb}
\usepackage{hyperref} 
\usepackage[round,semicolon]{natbib}

\newcommand{\comment}[1]{{\color{red}[\textsl{#1}]}}
\newcommand{\response}[1]{{\color{black}#1}}

\title{Response to reviews of ``An antibody-escape calculator for mutations to the SARS-CoV-2 receptor-binding domain'' for \textit{Virus Evolution}}
\author{}

\begin{document}
\maketitle

\emph{Below, the reviewer and editor comments {\color{blue} are in blue}, and may responses are in black.}

\color{blue}

\subsection*{Associate Editor: Adi Stern}

Apologies for the delayed response, reviewing during pandemic seems to have become tricky.
Both reviewers agree this is an excellent paper and a valuable resource.
Both suggest a series of slight changes and clarifications that all seem valuable and doable.
I agree with reviewer 2 that perhaps``estimator'' might be a better term than calculator to emphasize uncertainty. 
Finally, I agree with the authors discussion (and reviewer 2) on the caveat of these data being based on pre-variant escape data.
I do wonder how difficult it will be to implement newer versions that are based on future data with more complex variants?
Perhaps worth adding a few words on this as well to the discussion.

\response{Thanks to you and the reviewers for the helpful reviews. In the revised version, we have addressed all of the specific suggestions, including the three you mention above (see below for details).}

\subsection*{Reviewer: 1}

One of the key objectives for SARS-CoV-2 surveillance is to rapidly predict the impact that amino acid substitutions would have on important virus characteristics, including how well the spike receptor binding domain (RBD) will be recognized by therapeutic monoclonal antibodies and by polyclonal sera from subject following vaccination or infection.  The Bloom group has previously published a series of papers in which deep mutational scanning has been used to introduce all 20 amino acid substitutions at every site within the spike RBD and the impact on binding of monoclonal antibody and polyclonal antisera measured experimentally.  In this paper, Greaney et al. use this experimental data to develop an antibody-escape calculator to predict the functional impact of any combination of substitutions in the RBD on antibody binding.  They used data from three publications that measured virus neutralization of pseudoviruses carrying specific amino acid substitutions against convalescent or vaccine sera.  This calculator will be extremely helpful for the evaluation of emerging SARS-CoV-2 variants.

\comment{Thanks for the nice summary of the work.}

While the significance of the work is very high, the manuscript would be improved by addressing the following points.

For every monoclonal antibody, a single escape fraction value is calculated at each site by summarizing the impact of all 20 amino acid substitutions at the site as either the mean impact or the sum of the impacts to produce their escape maps.  When assessing the impact of a new mutation on binding escape, any mutation at that site appears to be treated equivalently.  An alternative approach would be to just use the specific escape fraction for the specific amino acid substitution at that site.  Thus, E484K would be treated differently from E484Q and E484A.  Indeed, one would expect that some substitutions would be more or less tolerated in terms of antibody binding.  The authors may have already evaluated this approach.  At a minimum, this point should be discussed in the manuscript.  But it might also be worthwhile to compare the performance of this more specific approach to the more generic approach implemented here.

It is a little difficult to know exactly how the different parameters are being calculated since in many cases they are just describe in the text.  For example, in the Methods section they state that “We also normalize the site-level escape metrics for each antibody to account for difference in strengths of antibody selection” and that “essentially this corresponds to scaling the escape value for each antibody so that a value of one corresponds to the larger of the maximum escape as a site or 20 times the median value across sites”.  It would be much easier to understand how these values are being calculated if the actual equations for each step in the workflow were provided.

There are several ways in which the binding score could be calculated, e.g., mean versus sum, specific substitutions versus generic substitutions, etc., but the authors only present the results from one approach.  It would be useful to present results from multiple configurations to determine if the results are robust to these configuration parameters.

The most important aspect of the paper is the correlation between the calculated binding score and the experimentally-measure fold change in virus neutralization as presented in Figure 3, which would provide some kind of validation of the computational method.  The authors state in the text that “the escape-calculator scores correlate quite well with the measured neutralization titers across all studies and cohorts”; “quite well” seems pretty subjective.  At a minimum, the strength of correlation should be quantified by using something like a Spearman’s rho metric which could then be used to calculate a t test statistic and p-value.  Indeed, these metrics could be used to compare the performance of the different configuration parameters.

It is not clear how to use the online calculator.  The instructions online refer to a Tweet chain but that provides a summary of how the calculations are performed, not on how to use the online tool.  It is not very intuitive and could use some simple explicit instructions on the page itself.

\subsection*{Reviewer: 2}

Greaney et al have developed a simple tool for estimating the impact of amino acid changes on antibody protection from SAR-CoV-2 variants. This is based on experimental data, much of which has been generated by these researchers. The text is clear and well written, the method is both sound and an easy to use implementation is provided. I only have some minor comments:

Please make it clear in the abstract that the experimental data is generated on the early pre-variant of concern data and so mutations may differ in consequence on new variants such as Omicron. This would seem to be quite a big caveat with this type of data.

Line 30, variants with reduced Ab neutralisation properties emerged before 2 years, e.g, Beta was designated in late 2020 in South Africa.

Line 41, briefly define variants of concern as there's ongoing confusion between these and just variants.

Line 54, the pop here at "complex black-box computational methods" is unwarranted and out of date relative to much contemporary machine leaning methods. Arguably simple residue based information from experiments is also a black-box as the mechanistic nuance of three-dimensional structural interactions are not really being accounted for. Maybe the authors could comment in the Discussion on why such a simple method works so well, e.g., linked to immunodominance of this spike region.

Two lines into Results, mutations strictly speaking occur at the nucleotide level and result in amino acid replacements. Clearly the authors know this and it fine to use mutations as short-hand for the residue changes but at least make the distinction clear the first time this is referred to.

Results, 3rd and 4th paragraph, let's not 'imagine' anything, please re-write.

Page 2, "illustrative" would be more accurate than "toy" example.

Finally, the authors should reflect on whether this is a "calculator" as opposed to an antibody-escape "estimator", i.e., capture the inaccuracy in the inferences a bit more clearly.

\response{This is a good suggestion, and we have changed the text to use ``estimator'' rather than ``calculator.''}

\color{black}
\bibliographystyle{mbe}
{\small
\bibliography{references.bib}
}

\end{document}

### 2022-03-06 paper/paper_original_submission.tex
set up to address reviewer comments

--- 2021-12-04 paper/paper.tex
@@
 A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
-Unfortunately, the rate at which new variants arise outstrips the speed with which these experiments can be performed.
+Unfortunately, the rate at which new variants arise outstrips the speed at which these experiments can be performed.
 
@@
 \subsection{The escape calculations correlate well with neutralization assays of human polyclonal sera against SARS-CoV-2 variants}
-For each set of RBD mutations, we can define a quantitative score that represents the polyclonal antibody binding that remains after making these mutations.
+For each set of mutated RBD sites, we can define a quantitative score that represents the polyclonal antibody binding that remains after mutating these sites.
 This score is defined using the same principle as the site-wise escape calculator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
@@
 \includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
-\caption{Correlation of calculated binding scores with experimentally measured fold changes in neutralization for SARS-CoV-2 variants and mutants.
-The data of \citet{lucas2021impact} was generated using live virus and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
+\caption{Correlation of calculated binding scores with experimentally measured fold changes in neutralization for SARS-CoV-2 variants and mutants (smaller values indicate worse neutralization).
+The data of \citet{lucas2021impact} was generated using authentic SARS-CoV-2 and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
 The data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with vaccine sera from Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively.
-The fold changes represent the geometric means over all subjects in each cohort.
+The fold changes are geometric means over all subjects in each cohort.
 An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html}.
@@
 We have described an escape calculator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
-The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus synergistic effects on antigenic escape.
+The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus additive effects on antibody binding.
 For instance, sites 417, 484, and 490 are all peaks of antibody escape.
-But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have synergistic effects since they generally escape different antibodies.
+But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have additive effects since they generally escape different antibodies.
 
@@
 There are caveats that should be kept in mind when using the escape calculator.
-First, it only assesses the antigenic effects of mutations, and not how they impact other properties such as viral transmissibility.
-Second, the calculator only considers sites in the RBD, and ignores mutations to other regions of spike.
-Third, the calculator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
-Fourth, the calculator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
+First, the calculator only considers sites in the RBD, and ignores mutations to other regions of spike.
+Second, the calculator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
+Third, the calculator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
 Finally, and in our minds most significantly, the calculator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).
@@
 Despite all these caveats, the escape calculator yields binding scores that correlate with experimentally measured neutralization titers.
-In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape calculator: variants of concern generally have combinations of mutations calculated to have synergistic effects on antigenic escape (eg, 417 and 484) rather than combinations calculated to have redundant effects (eg, 484 and 490).
+In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape calculator: variants of concern generally have combinations of mutations calculated to have additive effects on antibody escape (e.g., 417 and 484) rather than combinations calculated to have redundant effects (e.g., 484 and 490).
 We suspect the calculator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
@@
 Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the calculator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
-The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on sequences reported by Regeneron, Eli Lilly, or their collaborators~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
+The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on previously reported sequences~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
 The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).
@@
 The deep mutational scanning measures an escape fraction for each tolerated RBD mutation against each antibody, which represents an estimate of how completely that mutation escapes antibody binding~\citep{greaney2021complete}.
-We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions at each site.
+We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions across all tolerated mutations at each site.
 The results reported in this paper use the sums as the site-level metric, although the calculator has an option (\texttt{escape\_metric}) to use the mean instead.

### 2022-03-06 paper/paper.tex
change "calculator" -> "estimator"

@@
 
-\title{An antibody-escape calculator for mutations to the SARS-CoV-2 receptor-binding domain}
+\title{An antibody-escape estimator for mutations to the SARS-CoV-2 receptor-binding domain}
 
@@
 Unfortunately, direct experimental characterization of new viral variants lags their sequence-based identification. 
-Here we help address this challenge by aggregating deep mutational scanning data into an ``escape calculator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
-The calculator can be used to intuitively visualize how mutations impact polyclonal antibody recognition, and score the expected antigenic effect of combinations of mutations.
+Here we help address this challenge by aggregating deep mutational scanning data into an ``escape estimator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
+The estimator can be used to intuitively visualize how mutations impact polyclonal antibody recognition, and score the expected antigenic effect of combinations of mutations.
 These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
-An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
+An interactive version of the estimator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
 \end{abstract}
@@
 Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
-The resulting ``escape calculator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
-Importantly, the escape calculator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.
+The resulting ``escape estimator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
+Importantly, the escape estimator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.
 
@@
 
-\subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape calculator}
+\subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape estimator}
 The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
@@
 \caption{
-An escape calculator generated by aggregating deep mutational scanning for 33 neutralizing antibodies targeting the SARS-CoV-2 RBD.
+An escape estimator generated by aggregating deep mutational scanning for 33 neutralizing antibodies targeting the SARS-CoV-2 RBD.
 {\bf (A)} The blue line shows the extent of escape mediated by mutations at each site, as estimated by simply averaging the data for all the individual antibodies.
@@
 
-We recommend the reader explores the interactive escape calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, to perform calculations like those in Figure~\ref{fig:escape_calc} for arbitrary combinations of mutated RBD sites.
+We recommend the reader explores the interactive escape estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, to perform calculations like those in Figure~\ref{fig:escape_calc} for arbitrary combinations of mutated RBD sites.
 Such visual exploration of different combinations of mutations provides an intuitive sense of the antigenic structure of the RBD.
@@
 \subsection{The escape calculations correlate well with neutralization assays of human polyclonal sera against SARS-CoV-2 variants}
-For each set of RBD mutations, we can define a quantitative score that represents the polyclonal antibody binding that remains after making these mutations.
-This score is defined using the same principle as the site-wise escape calculator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
-This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).
+For each set of mutated RBD sites, we can define a quantitative score that represents the polyclonal antibody binding that remains after mutating these sites.
+This score is defined using the same principle as the site-wise escape estimator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
+This calculation is implemented in the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).
 
@@
 
-To test how these escape-calculator scores compare to experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
-One can imagine many reasons why the escape-calculator scores might differ from the real neutralization titers: the calculator only considers RBD mutations, the antibodies used by the calculator might not accurately reflect the real mix in polyclonal sera, etc.
-But despite all these potential caveats, the escape-calculator scores correlate quite well with the measured neutralization titers across all studies and cohorts (Figure~\ref{fig:neut_studies}).
-Therefore, the simple and intuitive approach used by the calculator seems to accurately reflect the dominant features of polyclonal antibody escape in the RBD.
+To test how these escape-estimator scores compare to experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
+One can imagine many reasons why the escape-estimator scores might differ from the real neutralization titers: the estimator only considers RBD mutations, the antibodies used by the estimator might not accurately reflect the real mix in polyclonal sera, etc.
+But despite all these potential caveats, the escape-estimator scores correlate quite well with the measured neutralization titers across all studies and cohorts (Figure~\ref{fig:neut_studies}).
+Therefore, the simple and intuitive approach used by the estimator seems to accurately reflect the dominant features of polyclonal antibody escape in the RBD.
 
-\subsection{The escape calculator suggests extensive antigenic change in the new Omicron variant}
-We applied the escape calculator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{ngs-sa2021update,deoliveira2021tweet}.
+\subsection{The escape estimator suggests extensive antigenic change in the new Omicron variant}
+We applied the escape estimator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{ngs-sa2021update,deoliveira2021tweet}.
 The calculated binding score for the Omicron variant is much lower than any other SARS-CoV-2 variants of concern, indicating extensive antibody escape (Figure~\ref{fig:Omicron}A).
@@
 The Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
-Exploration of the mutations using the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that mutations at sites 484, 446, and 417 are the biggest drivers of this antigenic change, although other mutations also contribute.
+Exploration of the mutations using the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that mutations at sites 484, 446, and 417 are the biggest drivers of this antigenic change, although other mutations also contribute.
 The residual peaks in the map suggests the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.
@@
 \section{Discussion}
-We have described an escape calculator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
-The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus synergistic effects on antigenic escape.
+We have described an escape estimator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
+The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus additive effects on antibody binding.
 For instance, sites 417, 484, and 490 are all peaks of antibody escape.
-But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have synergistic effects since they generally escape different antibodies.
+But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have additive effects since they generally escape different antibodies.
 
-Another key aspect of our approach is the interactive visual implementation of the escape calculator, so the user can interrogate the effects of mutations (or combinations of mutations) simply by clicking (or shift-clicking) on sites.
-This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the calculator is performing simple transformations directly on experimental data.
-We encourage interactive use of the calculator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
-However, we note that in addition to the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.
+Another key aspect of our approach is the interactive visual implementation of the escape estimator, so the user can interrogate the effects of mutations (or combinations of mutations) simply by clicking (or shift-clicking) on sites.
+This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the estimator is performing simple transformations directly on experimental data.
+We encourage interactive use of the estimator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
+However, we note that in addition to the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.
 
-There are caveats that should be kept in mind when using the escape calculator.
-First, it only assesses the antigenic effects of mutations, and not how they impact other properties such as viral transmissibility.
-Second, the calculator only considers sites in the RBD, and ignores mutations to other regions of spike.
-Third, the calculator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
-Fourth, the calculator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
-Finally, and in our minds most significantly, the calculator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).
+There are caveats that should be kept in mind when using the escape estimator.
+First, the estimator only considers sites in the RBD, and ignores mutations to other regions of spike.
+Second, the estimator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
+Third, the estimator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
+Finally, and in our minds most significantly, the estimator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).
 
-Despite all these caveats, the escape calculator yields binding scores that correlate with experimentally measured neutralization titers.
-In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape calculator: variants of concern generally have combinations of mutations calculated to have synergistic effects on antigenic escape (eg, 417 and 484) rather than combinations calculated to have redundant effects (eg, 484 and 490).
-We suspect the calculator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
+Despite all these caveats, the escape estimator yields binding scores that correlate with experimentally measured neutralization titers.
+In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape estimator: variants of concern generally have combinations of mutations calculated to have additive effects on antibody escape (e.g., 417 and 484) rather than combinations calculated to have redundant effects (e.g., 484 and 490).
+We suspect the estimator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
 
@@
 Additionally, differing exposure histories can leave individuals with different antibody specificities~\citep{cobey2017immune}, a process that is already starting to occur for SARS-CoV-2~\citep{greaney2021sars}.
-So in the future, it will be necessary to stratify the data used by the escape calculator by which RBD variant elicited the antibodies, and aggregate data for antibodies that reflect the sera in question.
-For this reason, we expect to continue adding to the data used by the escape calculator, and emphasize that it will change over time from the version described here, although we provide stable links to the current version in the Methods below.
+So in the future, it will be necessary to stratify the data used by the escape estimator by which RBD variant elicited the antibodies, and aggregate data for antibodies that reflect the sera in question.
+For this reason, we expect to continue adding to the data used by the escape estimator, and emphasize that it will change over time from the version described here, although we provide stable links to the current version in the Methods below.
 
@@
 \subsection{Code and data availability}
-The most up-to-date code and data used to implement the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps}, and the version described in this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/tree/bioRxiv_v1}.
+The most up-to-date code and data used to implement the escape estimator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps}, and the version described in this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/tree/bioRxiv_v1}.
 
-The data used by the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv}, and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv}.
+The data used by the escape estimator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv}, and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv}.
 
-A downloadable HTML version of the escape calculator is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/raw/main/docs/_includes/escape_calc_chart.html}, with the version described in this paper at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/docs/_includes/escape_calc_chart.html}.
+A downloadable HTML version of the escape estimator is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/raw/main/docs/_includes/escape_calc_chart.html}, with the version described in this paper at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/docs/_includes/escape_calc_chart.html}.
 
@@
 
-\subsection{Deep mutational scanning data used by the calculator}
-The experimental data used by the escape calculator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
+\subsection{Deep mutational scanning data used by the estimator}
+The experimental data used by the escape estimator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
 In total, these studies contain data for 36 monoclonal antibodies.
-Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the calculator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
-The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on sequences reported by Regeneron, Eli Lilly, or their collaborators~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
+Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the estimator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
+The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on previously reported sequences~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
 The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).
@@
 The deep mutational scanning measures an escape fraction for each tolerated RBD mutation against each antibody, which represents an estimate of how completely that mutation escapes antibody binding~\citep{greaney2021complete}.
-We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions at each site.
-The results reported in this paper use the sums as the site-level metric, although the calculator has an option (\texttt{escape\_metric}) to use the mean instead.
+We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions across all tolerated mutations at each site.
+The results reported in this paper use the sums as the site-level metric, although the estimator has an option (\texttt{escape\_metric}) to use the mean instead.
 We also normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
-The results reported in this paper use the normalized site-level metrics, although the calculator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.
+The results reported in this paper use the normalized site-level metrics, although the estimator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.
 
-\subsection{Calculation of the impact of mutations}
-The escape calculator determines the impact of mutating sites by calculating how much each antibody is escaped by mutations at each site, and adjusting its contribution to the overall polyclonal mix accordingly.
+\subsection{Estimation of the impact of mutations}
+The escape estimator determines the impact of mutating sites by calculating how much each antibody is escaped by mutations at each site, and adjusting its contribution to the overall polyclonal mix accordingly.
 
@@
 The variable $s$ represents how dramatically binding is lost for mutations at sites of escape that are not the strongest: larger values of $s$ means mutations at even moderate sites of escape reduce binding a lot.
-In this paper we report calculations with $s = 2$, although the calculator has an option (\texttt{mutation\_escape\_strength}) to choose other values.
+In this paper we report calculations with $s = 2$, although the estimator has an option (\texttt{mutation\_escape\_strength}) to choose other values.
 We then define the escape map after the mutations $\mathcal{M}$ as $\frac{1}{A}\sum_a x_{a,r} b_a\left(\mathcal{M}\right)$.
-The calculator shows the escape map with no mutations in gray, and that after mutations in blue.
+The estimator shows the escape map with no mutations in gray, and that after mutations in blue.
 
@@
 
-\subsection{Implementation of interactive calculator}
-The interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} is imlemented using Altair (\url{https://altair-viz.github.io/})~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.
+\subsection{Implementation of interactive estimator}
+The interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} is imlemented using Altair (\url{https://altair-viz.github.io/})~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.
 
-\subsection{Python module with batch-mode calculator}
-A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}, and has all the same options as the interactive calculator.
+\subsection{Python module with batch-mode estimator}
+A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}, and has all the same options as the interactive estimator.
 

### 2022-03-07 paper/reviewer_response.tex
explain site vs mutation level info

@@
 
-\comment{Thanks for the nice summary of the work.}
+\response{Thanks for the nice summary of the work.}
 
@@
 But it might also be worthwhile to compare the performance of this more specific approach to the more generic approach implemented here.
+
+\response{This is a good question, and we have elaborated on this point at both the beginning of the Results (to explain our approach) and in the Discussion (to mention this as a caveat).
+The major reason is that in practice mutation-level measurements tend to be a bit noisier in our data due (for instance) to specific mutations at a site sometimes being at low abundance in the deep mutational scanning libraries.
+Averaging over mutations at a site decreases the effects of this noise.
+For this reason, we think that the approximation of assuming mutations at a site have similar effects (which is often roughly true) is less bad than the decreased accuracy of the mutation-level compared to site-level measurements.
+A secondary reason is that using site-level measurements means that the calculations no longer have to be referred to a specific RBD sequence, which is useful if the immunodominance hierarchy remains the same even as the virus evolves and people are exposed to new variants (for instance, making antibodies against A484 rather than E484 after infection with Omicron).
+Overall, the new text should make clear that using the site-level measurements is an approximation for the reason the reviewer points out, but why we think it is overall a useful approximation.}
 

### 2022-03-07 paper/paper.tex
explain site vs mutation level info

@@
 Here we will work with these site-level escape maps.
+We use site-level information for two reasons: mutation-level measurements tend to be more noisy and averaging them for a site decreases this noise, and using site-level information makes the approach independent of the particular wildtype amino acid at a site (which is useful if we want to keep using the calculator as the RBD evolves).
+However, we note that site-level information ignores the possibility of different mutations at a site having different effects, and so mutation-level approaches could also become useful as the quality of experimental data improve.
 
@@
 Second, the estimator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
-Third, the estimator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
+Third, the estimator simply averages site-level escape measurements across antibodies, and ignores differences in the effects of different amino-acid mutations at the same site.
+Fourth, the estimator does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
 Finally, and in our minds most significantly, the estimator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).

### 2022-03-07 paper/paper.tex
equation for normalization

@@
 The results reported in this paper use the sums as the site-level metric, although the estimator has an option (\texttt{escape\_metric}) to use the mean instead.
-We also normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
-The results reported in this paper use the normalized site-level metrics, although the estimator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.
+
+We normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
+Specifically, the normalization is done so that for each antibody $a$, the maximum escape $x_{a,r}$ at any site $r$ is set so that $1 = \max\left(\max_r \left[x_{a,r}\right], 20\tilde{x}_{a,r} \right)$ where $\tilde{x}_{a,r}$ is the median of $x_{a,r}$ across sites $r$ for antibody $a$.
+The rationale for this normalization is that it usually scales the site-level escape metric so that the maximum value at a site for each antibody is one, but for very ``flat'' escape profiles where no site has more escape than 20 times the median, then the maximum value is smaller corresponding to no real peaks of escape for the antibody.
 

### 2022-03-07 paper/reviewer_response.tex
equation for normalization
add correlations and plot results for site mean as Fig S1

@@
 
+\response{This is a good suggestion, and we have added the exact equation used to perform the normalization in the relevant section of the Methods.}
+
 There are several ways in which the binding score could be calculated, e.g., mean versus sum, specific substitutions versus generic substitutions, etc., but the authors only present the results from one approach.
 It would be useful to present results from multiple configurations to determine if the results are robust to these configuration parameters.
+
+\response{
+This is a good point.
+It turns out that using the mean versus sum of mutations at a site gives extremely similar results.
+This can be seen graphically by going to the escape calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} and selecting the bottom metric to be \texttt{mean of mutations at site} rather than \texttt{sum of mutations at site}.
+To quantify this more formally, we have also added (as Supplementary Figure 1) a plot of the estimated escape versus experimental measurements using mean at a site in addition to the sum-at-a-site scores shown in Figure 3.
+The results are nearly identical: Pearson's R values of 0.91 vs 0.91, 0.72 vs 0.72, 0.88 vs 0.87, 0.82 vs 0.81, 0.76 vs 0.75, and 0.92 vs 0.91.
+We have added mention in the text of how the results are similar regardless of whether we use the mean or sum at a site.
+}
 
@@
 Indeed, these metrics could be used to compare the performance of the different configuration parameters.
+
+\response{
+We have added Pearson's R and the associated P-values in Figure 3.
+Five of the 6 correlations are significant, with $P < 0.05$.
+In addition, we have used correlation coefficients to compare the sum-of-mutations-at-a-site model to the mean-of-mutations-at-a-site model as described in the point immediately above.
+}
 

### 2022-03-07 paper/paper.tex
add correlations and plot results for site mean as Fig S1

@@
 The fold changes are geometric means over all subjects in each cohort.
+Labels at the top of each plot shown the Pearson's R and associated P-value.
 An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html}.
+Supplementary Figure S1 shows almost identical results are obtained if the binding scores are computed using the mean of mutations at a site rather than the sum.
 }
@@
 We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions across all tolerated mutations at each site.
-The results reported in this paper use the sums as the site-level metric, although the estimator has an option (\texttt{escape\_metric}) to use the mean instead.
+The results reported in this paper use the sums as the site-level metric, although the estimator has an option (\texttt{escape\_metric}) to use the mean instead (the results are almost identical: compare Fig~\ref{fig:neut_studies} versus Supplementary Figure S1.
 
@@
 We thank the Moyo and de Oliveira labs and other researchers in South Africa and Botswana for rapidly sharing information about the Omicron variant to enable the analyses in the last section of the results.
-This work was supported in part by the NIH/NIAID under R01AI141707 (to JDB) and T32AI083203 (to AJG).
+This work was supported in part by the NIH/NIAID under contract HHS75N93021C00015, and grants R01AI141707 (to JDB) and T32AI083203 (to AJG).
 The work was also supported by the Gates Foundation grant INV-004949 to JDB.
@@
 }
+
 \end{document}

### 2022-03-07 paper/supp.tex
add correlations and plot results for site mean as Fig S1

--- 2021-12-04 paper/paper.tex
@@
 \usepackage{stfloats}
-\usepackage[leftcaption]{sidecap}
-\sidecaptionvpos{figure}{t}
 
@@
 
-\title{An antibody-escape calculator for mutations to the SARS-CoV-2 receptor-binding domain}
-
-\author[*]{\Large Allison J. Greaney$^{1,2,3}$, Tyler N. Starr$^{1,4}$, Jesse D. Bloom$^{1,2,4}$}
-
-\affil[1]{Basic Sciences and Computational Biology, Fred Hutchinson Cancer Center
-
-} 
-\affil[2]{Department of Genome Sciences, University of Washington
-
-}
-\affil[3]{Medical Scientist Training Program, University of Washington
-
-}
-\affil[4]{Howard Hughes Medical Institute
-
-Seattle, WA, USA
-}
-
-\keywords{}
-
 \runningtitle{} % For use in the footer 
@@
 
-\begin{abstract}
-A key goal of SARS-CoV-2 surveillance is to rapidly identify viral variants with mutations that reduce neutralization by polyclonal antibodies elicited by vaccination or infection.
-Unfortunately, direct experimental characterization of new viral variants lags their sequence-based identification. 
-Here we help address this challenge by aggregating deep mutational scanning data into an ``escape calculator'' that estimates the antigenic effects of arbitrary combinations of mutations to the virus's spike receptor-binding domain (RBD).
-The calculator can be used to intuitively visualize how mutations impact polyclonal antibody recognition, and score the expected antigenic effect of combinations of mutations.
-These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
-An interactive version of the calculator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
-\end{abstract}
+\usepackage{xr}
+\externaldocument[]{paper}
 
@@
 
-\maketitle
-\thispagestyle{firststyle}
-%\marginmark
-\firstpagefootnote
+\onecolumn
+\renewcommand{\thepage}{S\arabic{page}}
+\setcounter{page}{1}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
 
-\correspondingauthoraffiliation{}{*Corresponding author: \href{mailto:jbloom@fredhutch.org}{jbloom@fredhutch.org}}
-\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
+\section{Supplementary Material}
 
-\lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
-This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have emerged only $\sim$2 years since the virus first started to spread in humans.
-A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
-Unfortunately, the rate at which new variants arise outstrips the speed with which these experiments can be performed.
+\begin{figure}[h!]
+\centering
+\includegraphics[width=\linewidth]{../results/neut_studies/neut_studies_mean.png}
+\caption{A version of Figure~\ref{fig:neut_studies} where the estimations are made using the mean effects of tolerated mutations at a site rather than the total (sum) of the effects of tolerated mutations at a site.
+The results are almost identical.
+}
+\label{suppfig:coverage}
+\end{figure}
 
-A partial solution is to use deep mutational scanning experiments to \emph{prospectively} measure how viral mutations impact antibody binding or neutralization.
-Deep mutational scanning can systematically measure the antigenic impacts of all possible amino-acid mutations in key regions of spike on monoclonal antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
-However, SARS-CoV-2 variants of concern typically have multiple mutations, and it is not feasible to experimentally characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.
-
-Here we take a step towards addressing this challenge by aggregating deep mutational scanning data across many antibodies to assess the impacts of mutations in the spike receptor-binding domain (RBD), which is the primary target of neutralizing antibodies to SARS-CoV-2~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high}.
-The resulting ``escape calculator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
-Importantly, the escape calculator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.
-
-\section{Results}
-
-\subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
-A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}.
-This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
-Here we will work with these site-level escape maps.
-
-As a small example to illustrate the principle behind our approach, Figure~\ref{fig:mini_example}A shows previously reported measurements~\citep{starr2021prospective, starr2021complete} of how mutations to each RBD site affect binding by three monoclonal antibodies: LY-CoV016 (etesevimab), LY-CoV555 (bamlanivimab), and REGN10987 (imdevimab).
-Each antibody targets a different epitope on the RBD: LY-CoV016 targets the class 1 epitope, LY-CoV555 the class 2 epitope, and REGN10987 the class 3 epitope~\citep{barnes2020sars,greaney2021mapping}.
-Because the antibodies have distinct epitopes, they are escaped by largely distinct sets of mutations: LY-CoV016 is most strongly escaped by mutations at site 417, LY-CoV555 at site 484, and REGN10987 at sites 444--446 (Figure~\ref{fig:mini_example}A).
-
-Now imagine a polyclonal antibody mix of these three antibodies combined at equal potencies.
-We can generate an escape map for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
-Because this polyclonal escape map is the average of the monoclonal antibody maps, its largest peaks are at the sites of strongest escape for each individual antibody: 417, 484, and 444--446.
-
-Next imagine removing one antibody from the mix by mutating its epitope.
-Figure~\ref{fig:mini_example}B shows the resulting escape map if LY-CoV555 is ablated, as would occur if site 484 was mutated.
-The thick black line for the antibody mix no longer has peaks at 484 and other sites targeted by LY-CoV555, such as 490.
-Therefore, in this hypothetical polyclonal antibody mix, escape at sites 484 and 490 is correlated since both sites are targeted by the same antibody.
-However, the polyclonal mix's escape map at sites 417 and 460 is unaffected by mutations that escape LY-CoV555, since they are targeted by a different antibody, LY-CoV016.
-But if we also ablate LY-CoV016 (such as by mutating site 417), then the peaks at 417 and 460 also disappear, and the remaining peaks are at sites targeted by REGN10987, such as 444--446 (Figure~\ref{fig:mini_example}C).
-Of course, if REGN10987 was also ablated such as by mutating site 446, then the polyclonal antibody mix would have no remaining activity.
-This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.
-
-\begin{SCfigure*}[][]
-\includegraphics[width=0.71\textwidth]{figures/mini_example/mini_example.pdf}
-
-\caption{
-Escape map for a hypothetical polyclonal mix consisting of an equipotent mixture of three monoclonal antibodies targeting distinct epitopes on the SARS-CoV-2 RBD.
-{\bf (A)} Experimentally measured escape maps for three antibodies, and the mean of these maps (thick black line).
-Each point on the x-axis represents a site in the RBD, and the y-axis represents the total measured escape by all mutations at that site scaled so the maximum for each antibody is one.
-{\bf (B)} Escape map if the contribution of antibody LY-CoV555 is ablated.
-{\bf (C)} Escape map if the contributions of antibodies LY-CoV555 and LY-CoV016 are ablated.
-An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.}
-\label{fig:mini_example}
-\end{SCfigure*}
-
-\subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape calculator}
-The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
-To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing antibodies elicited by SARS-CoV-2.
-These antibodies were isolated from a variety of patient cohorts within the first year of the pandemic (see Methods for details).
-An assumption of the analysis that follows is that an equipotent mixture of these 33 antibodies represents the neutralizing activity of human sera; we emphasize that this assumption is imperfect since in reality the antibodies were chosen for prior study for a variety of ad hoc reasons.
-The escape maps for all the individual antibodies can be interactively interrogated at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/}.
-
-\begin{SCfigure*}[][]
-\includegraphics[width=0.8\textwidth]{figures/escape_calc/escape_calc.pdf}
-\caption{
-An escape calculator generated by aggregating deep mutational scanning for 33 neutralizing antibodies targeting the SARS-CoV-2 RBD.
-{\bf (A)} The blue line shows the extent of escape mediated by mutations at each site, as estimated by simply averaging the data for all the individual antibodies.
-{\bf (B)} The blue line shows escape map after a mutation to site 484 (red point) ablates recognition by antibodies strongly targeting that site, while the gray line shows the original escape map in the absence of any mutations.
-{\bf (C)} The escape map after mutating sites 417, 484, and 501 (the three RBD sites mutated in the Beta variant).
-An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}.}
-\label{fig:escape_calc}
-\end{SCfigure*}
-
-The overall polyclonal escape map generated by averaging the experimental data for all 33 antibodies is in Figure \ref{fig:escape_calc}A.
-As in the toy three-antibody example in the previous section, there are peaks at  sites 417, 484, and 444--446.
-However, the peak at 484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are especially common in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
-In addition, there are smaller peaks at a variety of other sites, reflecting the fact that each antibody has a somewhat idiosyncratic epitope (Figure \ref{fig:escape_calc}A).
-
-We can follow the principle outlined in the toy example in the previous section to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
-Specifically, we reduce the contribution of each antibody by an amount that scales with how strongly that antibody targets each mutated site (see Methods for details).
-For instance, the blue lines in Figure~\ref{fig:escape_calc}B show the polyclonal escape map after mutating site 484.
-Mutating site 484 obviously drops the contribution of that site, but it also decreases the contribution of other sites such as 490 that are commonly targeted by antibodies with epitopes that include site 484.
-In contrast, mutating site 484 has minimal effect on the polyclonal escape map at sites like 417 or 444--446, since those sites are generally targeted by antibodies that are unaffected by mutations at site 484.
-
-We can also calculate the expected effects of compound mutations.
-Figure~\ref{fig:escape_calc}C shows the polyclonal escape map after mutating all three RBD sites that are changed in the Beta variant (sites 417, 484, and 501).
-This polyclonal escape map has lost contributions not only from the mutated sites, but also sites that form common epitopes with 417 or 484 (e.g., sites 455, 456, 486, and 490).
-However, the escape map still has major contributions from antibodies targeting sites like 444--446, since such antibodies are generally unaffected by mutations at sites 417, 484, or 501.
-
-We recommend the reader explores the interactive escape calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, to perform calculations like those in Figure~\ref{fig:escape_calc} for arbitrary combinations of mutated RBD sites.
-Such visual exploration of different combinations of mutations provides an intuitive sense of the antigenic structure of the RBD.
-
-\subsection{The escape calculations correlate well with neutralization assays on SARS-CoV-2 variants}
-For each set of RBD mutations, we can define a quantitative score that represents the polyclonal antibody binding that remains after making these mutations.
-This score is defined using the same principle as the site-wise escape calculator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
-This calculation is implemented in the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).
-
-\begin{SCfigure*}[][]
-\includegraphics[width=0.7\textwidth]{../results/neut_studies/neut_studies.png}
-\caption{Correlation of escape scores from the calculator with experimentally measured fold changes in neutralization IC50 values for SARS-CoV-2 variants and mutants.
-The experimental data of \citet{lucas2021impact} was generated using live virus and sera from vaccinated individuals who were (top) or were not (bottom) previously infected with SARS-CoV-2.
-The experimental data of \citet{uriu2021neutralization} and \citet{wang2021antibody} was generated using pseudovirus against convalescent (top) or vaccine (bottom) sera, with the vaccine sera from the Pfizer BNT162b2 or Moderna mRNA-1273 vaccines, respectively.
-The plotted fold changes in neutralization represent the geometric means over all subjects in each cohort.
-}
-\label{fig:neut_studies}
-\end{SCfigure*}
-
-To test how these escape-calculator scores compare to experimentally measured neutralization titers, we collated neutralization data from three previously published studies~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}, each of which characterized sera from two patient cohorts against a variety of SARS-CoV-2 variants and mutants.
-One can imagine many reasons why the escape-calculator scores might differ from the real neutralization titers: the calculator only considers RBD mutations, the antibodies used by the calculator might not accurately reflect the real mix in polyclonal sera, etc.
-But despite all these potential caveats, the escape-calculator scores correlate quite well with the measured neutralization titers across all studies and cohorts (Figure~\ref{fig:neut_studies}).
-Therefore, the simple and intuitive approach used by the calculator seems to accurately reflect the dominant features of polyclonal antibody escape in the RBD.
-
-\subsection{The escape calculator suggests extensive antigenic change in the new Omicron variant}
-We applied the escape calculator to the recently reported Omicron variant, which has 15 mutated sites in its RBD~\citep{ngs-sa2021update,deoliveira2021tweet}.
-The calculated binding score for the Omicron variant is much lower than any other SARS-CoV-2 variants of concern, indicating extensive antibody escape (Figure~\ref{fig:Omicron}A).
-The Omicron variant's calculated score is roughly equivalent to that of a polymutant spike (PMS20) that was artificially engineered in a pseudovirus by \citet{schmidt2021high} to maximize escape from polyclonal serum antibodies.
-For comparison, \citet{schmidt2021high} measured that neutralization titers against this artificial PMS20 spike were reduced by $\sim$20- to $\sim$80-fold for sera from various cohorts of vaccinated and infected individuals.
-
-\begin{figure*}
-\includegraphics[width=\linewidth]{figures/Omicron/Omicron.pdf}
-\caption{Escape calculations for the Omicron variant.
-{\bf (A)} The calculated binding scores for SARS-CoV-2 variants and the artificial polymutant spike (PMS20) generated by \citet{schmidt2021high}.
-Scores of one indicate no mutations affect binding, and scores of zero indicate no antibody binding remains.
-{\bf (B)} The calculated escape map for the Omicron variant's RBD (blue) compared to an unmutated RBD (gray), with sites of mutations in the Omicron variant in red.
-The mutated RBD sites for each variant are in Table~\ref{tab:variants}.
-}
-\label{fig:Omicron}
-\end{figure*}
-
-The site-level escape map for the Omicron variant's RBD is shown in Figure~\ref{fig:Omicron}B.
-The Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
-Exploration of the mutations using the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that mutations at sites 484, 446, and 417 are the biggest drivers of this antigenic change, although other mutations also contribute.
-The residual peaks in the map suggests the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.
-
-\section{Discussion}
-We have described an escape calculator that uses experimental data for 33 monoclonal antibodies to estimate the antigenic effects of arbitrary combinations of mutations to the SARS-CoV-2 RBD.
-The key insight is to aggregate data for individual antibodies to define both which RBD sites are antigenically important, and which combinations of mutations have redundant versus synergistic effects on antigenic escape.
-For instance, sites 417, 484, and 490 are all peaks of antibody escape.
-But mutations at 484 and 490 have redundant effects since they generally escape the same antibodies, whereas mutations at 484 and 417 have synergistic effects since they generally escape different antibodies.
-
-Another key aspect of our approach is the interactive visual implementation of the escape calculator, so the user can interrogate the effects of mutations (or combinations of mutations) simply by clicking (or shift-clicking) on sites.
-This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the calculator is performing simple transformations directly on experimental data.
-We encourage interactive use of the calculator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
-However, we note that in addition to the interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/bindingcalculator.py}.
-
-There are caveats that should be kept in mind when using the escape calculator.
-First, it only assesses the antigenic effects of mutations, and not how they impact other properties such as viral transmissibility.
-Second, the calculator only considers sites in the RBD, and ignores mutations to other regions of spike.
-Third, the calculator assumes the neutralizing activity of human polyclonal serum is represented by an equipotent mix of the monoclonal antibodies that happen to have been previously characterized by deep mutational scanning.
-Fourth, the calculator simply averages site-level escape measurements across antibodies, and does not yet implement a real biophysical model of the combined activity of multiple antibodies~\citep{einav2020two}.
-Finally, and in our minds most significantly, the calculator estimates the impact of mutations in reference to antibodies targeted to the early Wuhan-Hu-1 RBD---an approach that is currently reasonable, but will become problematic as human exposure and vaccination histories diversify in the years to come (see last paragraph).
-
-Despite all these caveats, the escape calculator yields binding scores that correlate with experimentally measured neutralization titers.
-In addition, the actual antigenic evolution of SARS-CoV-2 seems to follow the principles captured by the escape calculator: variants of concern generally have combinations of mutations calculated to have synergistic effects on antigenic escape (eg, 417 and 484) rather than combinations calculated to have redundant effects (eg, 484 and 490).
-We suspect the calculator works well because the RBD is the dominant target of neutralizing activity~\citep{piccoli2020mapping,greaney2021comprehensive,schmidt2021high} and the human antibody-response to the early Wuhan-Hu-1 RBD shares broad commonalities across individuals~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
-
-However, the situation will become more complex over time.
-Currently, most humans with antibodies to SARS-CoV-2 have been exposed to a RBD antigen that is identical or very similar to that of the early Wuhan-Hu-1 strain.
-Therefore, antigenic studies can reasonably define mutations in reference to that RBD, since it is what the antibodies target.
-But as humans are exposed to more diverged RBD variants, it will become difficult to determine what reference to use to define antigenic mutations, since different individuals will have antibodies targeting different RBDs.
-Additionally, differing exposure histories can leave individuals with different antibody specificities~\citep{cobey2017immune}, a process that is already starting to occur for SARS-CoV-2~\citep{greaney2021sars}.
-So in the future, it will be necessary to stratify the data used by the escape calculator by which RBD variant elicited the antibodies, and aggregate data for antibodies that reflect the sera in question.
-For this reason, we expect to continue adding to the data used by the escape calculator, and emphasize that it will change over time from the version described here, although we provide stable links to the current version in the Methods below.
-
-{\small
-
-\section{Methods}
-\subsection{Code and data availability}
-The most up-to-date code and data used to implement the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps}, and the version described in this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/tree/bioRxiv_v1}.
-
-The data used by the escape calculator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv}, and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv}.
-
-A downloadable HTML version of the escape calculator is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/raw/main/docs/_includes/escape_calc_chart.html}, with the version described in this paper at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/docs/_includes/escape_calc_chart.html}.
-
-\subsection{Deep mutational scanning data used by the calculator}
-The experimental data used by the escape calculator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
-In total, these studies contain data for 36 monoclonal antibodies.
-Three of these antibodies (CR3022, S304, and S309) were elicited by infection with SARS-CoV-1 and so are excluded from the datasets used for the calculations in this paper, although the calculator has an option (\texttt{eliciting\_virus}) that allows optional inclusion of these antibodies.
-The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and where initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on sequences reported by Regeneron, Eli Lilly, or their collaborators~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
-The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).
-
-The deep mutational scanning measures an escape fraction for each tolerated RBD mutation against each antibody, which represents an estimate of how completely that mutation escapes antibody binding~\citep{greaney2021complete}.
-We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions at each site.
-The results reported in this paper use the sums as the site-level metric, although the calculator has an option (\texttt{escape\_metric}) to use the mean instead.
-We also normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
-The results reported in this paper use the normalized site-level metrics, although the calculator has an option (\texttt{escape\_values\_normalized}) to also use the non-normalized values.
-
-\subsection{Calculation of the impact of mutations}
-The escape calculator determines the impact of mutating sites by calculating how much each antibody is escaped by mutations at each site, and adjusting its contribution to the overall polyclonal mix accordingly.
-
-Specifically, for each antibody $a$ we have a deep mutational scanning measurement $x_{a,r}$ of how much mutating $r$ escapes that antibody.
-In the absence of any mutations, the overall escape map shown for the polyclonal mix is simply the mean over all antibodies, $\frac{1}{A}\sum_a x_{a,r}$ where $A$ is the number of antibodies.
-
-Let $\mathcal{M}$ be the set of sites that are mutated.
-Then for each antibody we compute the binding retained as $b_a\left(\mathcal{M}\right) = \left(\prod\limits_{r \in \mathcal{M}} \frac{\max_{r'}\left(x_{a,r'}\right) - x_{a,r}}{\max_{r'}\left(x_{a,r'}\right)}\right)^s$.
-Essentially, this equation means that if the RBD is mutated at a strong site of escape for an antibody $a$, much if the binding is lost (if it is the strongest site of escape, all binding is lost).
-The variable $s$ represents how dramatically binding is lost for mutations at sites of escape that are not the strongest: larger values of $s$ means mutations at even moderate sites of escape reduce binding a lot.
-In this paper we report calculations with $s = 2$, although the calculator has an option (\texttt{mutation\_escape\_strength}) to choose other values.
-We then define the escape map after the mutations $\mathcal{M}$ as $\frac{1}{A}\sum_a x_{a,r} b_a\left(\mathcal{M}\right)$.
-The calculator shows the escape map with no mutations in gray, and that after mutations in blue.
-
-The overall antibody binding scores represent the fraction of antibodies that still bind, and are calculated simply as $\frac{1}{A}\sum_a b_a\left(\mathcal{M}\right)$.
-
-\subsection{Implementation of interactive calculator}
-The interactive calculator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} is imlemented using Altair (\url{https://altair-viz.github.io/})~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.
-
-\subsection{Python module with batch-mode calculator}
-A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/bindingcalculator.py}, and has all the same options as the interactive calculator.
-
-\subsection{Compilation of neutralization titers from the literature}
-For Figure~\ref{fig:neut_studies}, we compiled neutralization data from three published studies on SARS-CoV-2 variants and mutants~\citep{lucas2021impact,uriu2021neutralization,wang2021antibody}.
-For each study cohort, we computed the geometric mean fold change in neutralization titer over all subjects.
-The numerical compiled data are at \url{https://github.com/jbloomlab/RBD_escape_calculator_paper/tree/main/results/neut_studies}.
-
-\subsection{Mutations in SARS-CoV-2 variants}
-For Figure~\ref{fig:Omicron}, the definitions of which RBD sites are mutated in each variant are shown in Table~\ref{tab:variants}.
-\begin{table}[h]
-{\footnotesize
-\begin{tabular}{ll}
-\toprule
-   variant &                                           mutated RBD sites \\
-\midrule
-Wuhan-Hu-1 &                                                             \\
-     Alpha &                                                         501 \\
-      Beta &                                                 417 484 501 \\
-     Gamma &                                                 417 484 501 \\
-     Delta &                                                     452 478 \\
-    Lambda &                                                     452 490 \\
-        Mu &                                                 346 484 501 \\
-   Omicron & 339 371 373 375 417 440 446 477 478 484 493 496 498 501 505 \\
-     PMS20 &                             346 417 440 445 455 475 484 501 \\
-\bottomrule
-\end{tabular}
-}
-\caption{Mutated RBD sites in SARS-CoV-2 variants.}
-\label{tab:variants}
-\end{table}
-
-\section{Acknowledgments}
-We thank collaborators in the Crowe lab, Nussenzweig lab, Bjorkman lab, and at Vir Biotechnology for sharing the antibodies that were used in prior published studies to generate the deep mutational scanning data aggregated here.
-We thank the Moyo and de Oliveira labs and other researchers in South Africa and Botswana for rapidly sharing information about the Omicron variant to enable the analyses in the last section of the results.
-This work was supported in part by the NIH/NIAID under R01AI141707 (to JDB) and T32AI083203 (to AJG).
-The work was also supported by the Gates Foundation grant INV-004949 to JDB.
-TNS is a Howard Hughes Medical Institute Fellow of the Damon Runyon Cancer Research Foundation.
-JDB is an Investigator of the Howard Hughes Medical Institute.
-
-\section{Competing interests}
-JDB consults for Moderna, Flagship Labs 77 and Oncorus.
-JDB is an inventor on a Fred Hutch licensed patents related to deep mutational scanning of viral proteins.
-
-\bibliography{references}
-}
 \end{document}

### 2022-03-07 paper/reviewer_response.tex
instructions for online calculator
responses to reviewer 2

@@
 
+\response{We have added more instructions on the online page.}
+
 \subsection*{Reviewer: 2}
@@
 
+\response{Thanks for the positive assessment.}
+
 Please make it clear in the abstract that the experimental data is generated on the early pre-variant of concern data and so mutations may differ in consequence on new variants such as Omicron.
@@
 
+\response{This is a good point, and we have added the following sentence to the Abstract: \emph{Currently the calculator uses primarily data for antibodies elicited by Wuhan-Hu-1-like vaccination or infection, and so is expected to work best for calculating escape from such immunity for mutations relative to early SARS-CoV-2 strains.}.
+We do note that we anticipate updating as new data become available.}
+
 Line 30, variants with reduced Ab neutralisation properties emerged before 2 years, e.g, Beta was designated in late 2020 in South Africa.
 
+\response{Good point, we have clarified to indicate that new variants emerged within a year.}
+
 Line 41, briefly define variants of concern as there's ongoing confusion between these and just variants.
+
+\response{We have clarified that variants of concern are variants with reduced immune recognition, enhanced transmissibility, or increased virulence.}
 
@@
 
+\response{We have removed the word ``black box'' and clarified that we just mean that the computations can be interpreted visually using the online calculator.
+We have also elaborated on immunodominance.}
+
 Two lines into Results, mutations strictly speaking occur at the nucleotide level and result in amino acid replacements.
@@
 
+\response{We have clarified this directly: \emph{... where we here consider mutations at the level of amino-acid replacements rather than in terms of the underlying nucleotide changes.}}
+
 Results, 3rd and 4th paragraph, let's not 'imagine' anything, please re-write.
 
+\response{We have re-worded to make clear that we mean to consider a hypothetical antibody mix.}
+
 Page 2, "illustrative" would be more accurate than "toy" example.
+
+\response{We have changed ``toy'' to ``illustrative'' as suggested.}
 

### 2022-03-07 paper/paper.tex
responses to reviewer 2

@@
 An interactive version of the estimator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
+Currently the calculator uses primarily data for antibodies elicited by Wuhan-Hu-1-like vaccination or infection, and so is expected to work best for calculating escape from such immunity for mutations relative to early SARS-CoV-2 strains.
 \end{abstract}
@@
 \lettrine[lines=2]{\color{color2}H}{}uman coronaviruses undergo antigenic evolution that erodes antibody-based neutralization~\citep{eguia2021human,kistler2021evidence}.
-This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization have emerged only $\sim$2 years since the virus first started to spread in humans.
+This antigenic evolution is already apparent for SARS-CoV-2, as new viral variants with reduced antibody neutralization emerged within a year of when the virus first started to spread in humans.
 A tremendous amount of experimental effort has been expended to characterize these SARS-CoV-2 variants in neutralization assays~\citep{wang2021antibody,uriu2021neutralization,lucas2021impact}.
@@
 Deep mutational scanning can systematically measure the antigenic impacts of all possible amino-acid mutations in key regions of spike on monoclonal antibodies~\citep{starr2021prospective,greaney2021complete} or sera~\citep{greaney2021comprehensive}.
-However, SARS-CoV-2 variants of concern typically have multiple mutations, and it is not feasible to experimentally characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.
+However, SARS-CoV-2 variants of concern (variants with reduced immune recognition, enhanced transmissibility, or increased virulence) typically have multiple mutations, and it is not feasible to experimentally characterize all combinations of mutations even via high-throughput approaches such as deep mutational scanning.
 
@@
 The resulting ``escape estimator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
-Importantly, the escape estimator is based on simple transformations of direct experimental measurements, and does not involve complex black-box computational methods.
+Importantly, the escape estimator is based on simple transformations of direct experimental measurements, and and so its calculations can be intuitively visualized using the interactive interface we provide.
 
@@
 \subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
-A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}.
+A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}, where we here consider mutations at the level of amino-acid replacements rather than in terms of the underlying nucleotide changes.
 This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
@@
 
-Now imagine a polyclonal antibody mix of these three antibodies combined at equal potencies.
+Now consider a hypothetical polyclonal antibody mix of these three antibodies combined at equal potencies.
 We can generate an escape map for this hypothetical antibody mix simply by averaging the experimentally measured escape maps for the three individual antibodies, yielding the thick black line in Figure~\ref{fig:mini_example}A.
@@
 
-Next imagine removing one antibody from the mix by mutating its epitope.
+Next consider removing one antibody from the hypothetical mix by mutating its epitope.
 Figure~\ref{fig:mini_example}B shows the resulting escape map if LY-CoV555 is ablated, as would occur if site 484 was mutated.
@@
 \subsection{Aggregating deep mutational scanning data for 33 human antibodies yields a realistic escape estimator}
-The toy example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
+The illustrative example in the previous section illustrates how experimental data for individual antibodies can be combined to yield an escape map for a hypothetical polyclonal antibody mix.
 To create an escape map for an antibody mix that more realistically represents actual human sera, we aggregated previously generated deep mutational scanning data for 33 neutralizing antibodies elicited by SARS-CoV-2.
@@
 The overall polyclonal escape map generated by averaging the experimental data for all 33 antibodies is in Figure \ref{fig:escape_calc}A.
-As in the toy three-antibody example in the previous section, there are peaks at  sites 417, 484, and 444--446.
+As in the illustrative three-antibody example in the previous section, there are peaks at  sites 417, 484, and 444--446.
 However, the peak at 484 is now larger than any other peak, reflecting the fact that antibodies targeting the class 2 epitope containing E484 are especially common in the human antibody response to early SARS-CoV-2 strains~\citep{yuan2020structural,robbiani2020convergent,greaney2021comprehensive,greaney2021mapping,chen2021convergent}.
@@
 
-We can follow the principle outlined in the toy example in the previous section to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
+We can follow the principle outlined in the illustrative example in the previous section to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
 Specifically, we reduce the contribution of each antibody by an amount that scales with how strongly that antibody targets each mutated site (see Methods for details).
@@
 This interactivity provides an intuitive understanding of the antigenic structure of the RBD, and shows how the estimator is performing simple transformations directly on experimental data.
-We encourage interactive use of the estimator so it acts as an aid to augment human interpretation, rather than a black-box computational method.
+We encourage interactive use of the estimator so it acts as a visual aid to augment human interpretation.
 However, we note that in addition to the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.

### 2022-03-07 paper/reviewer_response.tex
minor fixes

@@
 
-\response{This is a good point, and we have added the following sentence to the Abstract: \emph{Currently the calculator uses primarily data for antibodies elicited by Wuhan-Hu-1-like vaccination or infection, and so is expected to work best for calculating escape from such immunity for mutations relative to early SARS-CoV-2 strains.}.
-We do note that we anticipate updating as new data become available.}
+\response{This is a good point, and we have added the following sentence to the Abstract: \emph{Currently the calculator uses primarily data for antibodies elicited by Wuhan-Hu-1-like vaccination or infection, and so is expected to work best for calculating escape from such immunity for mutations relative to early SARS-CoV-2 strains.} We do note that we anticipate updating as new data become available.}
 

### 2022-03-07 paper/paper.tex
minor fixes

@@
 
-We can follow the principle outlined in the illustrative example in the previous section to calculate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
+We can follow the principle outlined in the illustrative example in the previous section to estimate the expected polyclonal escape map \emph{after} mutating sites in the RBD.
 Specifically, we reduce the contribution of each antibody by an amount that scales with how strongly that antibody targets each mutated site (see Methods for details).

### 2022-03-08 paper/paper.tex
a few minor tiny edits from Tyler and Allie

@@
 The resulting ``escape estimator'' enables qualitative visualization and quantitative scoring of the antigenic effects of arbitrary combinations of mutations.
-Importantly, the escape estimator is based on simple transformations of direct experimental measurements, and and so its calculations can be intuitively visualized using the interactive interface we provide.
+Importantly, the escape estimator is based on simple transformations of direct experimental measurements, and so its calculations can be intuitively visualized using the interactive interface we provide.
 
@@
 \subsection{Combining monoclonal antibody escape maps reveals correlated and independent viral antigenic mutations}
-A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}, where we here consider mutations at the level of amino-acid replacements rather than in terms of the underlying nucleotide changes.
+A deep mutational scanning experiment can measure how all single amino-acid mutations to the SARS-CoV-2 RBD affect binding by a monoclonal antibody~\citep{greaney2021complete}.
 This mutation-level information can be summarized for each RBD site, such as by taking the mean or sum of mutation-level effects at a site.
@@
 We summarize the mutation-level escape fractions into site-level measurements in two ways: taking the sum of the mutation escape fractions at each site, or taking the mean of the mutation escape fractions across all tolerated mutations at each site.
-The results reported in this paper use the sums as the site-level metric, although the estimator has an option (\texttt{escape\_metric}) to use the mean instead (the results are almost identical: compare Fig~\ref{fig:neut_studies} versus Supplementary Figure S1.
+The results reported in this paper use the sums as the site-level metric, although the estimator has an option (\texttt{escape\_metric}) to use the mean instead (the results are almost identical: compare Figure~\ref{fig:neut_studies} versus Supplementary Figure S1.
 

### 2022-03-11 paper/paper.tex
add "last accessed" to all URL refs

@@
 These scores correlate with neutralization assays performed on SARS-CoV-2 variants, and emphasize the ominous antigenic properties of the recently described Omicron variant.
-An interactive version of the estimator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and we provide a Python module for batch processing.
+An interactive version of the estimator is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} (last accessed March-11-2022), and we provide a Python module for batch processing.
 Currently the calculator uses primarily data for antibodies elicited by Wuhan-Hu-1-like vaccination or infection, and so is expected to work best for calculating escape from such immunity for mutations relative to early SARS-CoV-2 strains.
@@
 Of course, if REGN10987 was also ablated such as by mutating site 446, then the polyclonal antibody mix would have no remaining activity.
-This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.
+This and other scenarios can be explored using the interactive version of Figure~\ref{fig:mini_example} at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/} (last accessed March-11-2022).
 
@@
 {\bf (C)} Escape map if the contributions of antibodies LY-CoV555 and LY-CoV016 are ablated.
-An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/}.}
+An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/mini-example-escape-calc/} (last accessed March-11-2022).}
 \label{fig:mini_example}
@@
 An assumption of the analysis that follows is that an equipotent mixture of these 33 antibodies represents the neutralizing activity of human sera; we emphasize that this assumption is imperfect since in reality the antibodies were chosen for prior study for a variety of ad hoc reasons.
-The escape maps for all the individual antibodies can be interactively interrogated at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/}.
+The escape maps for all the individual antibodies can be interactively interrogated at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} (last accessed March-11-2022).
 
@@
 {\bf (C)} The escape map after mutating sites 417, 484, and 501 (the three RBD sites mutated in the Beta variant).
-An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}.}
+An interactive version of this figure is at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} (last accessed March-11-2022).}
 \label{fig:escape_calc}
@@
 
-We recommend the reader explores the interactive escape estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, to perform calculations like those in Figure~\ref{fig:escape_calc} for arbitrary combinations of mutated RBD sites.
+We recommend the reader explores the interactive escape estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} (last accessed March-11-2022), to perform calculations like those in Figure~\ref{fig:escape_calc} for arbitrary combinations of mutated RBD sites.
 Such visual exploration of different combinations of mutations provides an intuitive sense of the antigenic structure of the RBD.
@@
 This score is defined using the same principle as the site-wise escape estimator in the previous section: we reduce the contribution of each antibody by an amount that scales with how strongly it is escaped by each mutated site, and define the overall score as the fraction of all antibody contributions that remain (see Methods for details).
-This calculation is implemented in the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).
+This calculation is implemented in the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} (last accessed March-11-2022), and returns a score that ranges from one (no mutations affect binding of any antibodies) to zero (all antibodies fully escaped).
 
@@
 Labels at the top of each plot shown the Pearson's R and associated P-value.
-An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html}.
+An interactive version of this figure that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/neut_studies.html} (last accessed March-11-2022).
 Supplementary Figure S1 shows almost identical results are obtained if the binding scores are computed using the mean of mutations at a site rather than the sum.
@@
 Scores of one indicate no mutations affect binding, and scores of zero indicate no antibody binding remains.
-An interactive version of this plot that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/variants.html}.
+An interactive version of this plot that allows mousing over points to see details is at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/variants.html} (last accessed March-11-2022).
 {\bf (B)} The calculated escape map for the Omicron variant's RBD (blue) compared to an unmutated RBD (gray), with sites of mutations in the Omicron variant in red.
@@
 The Omicron RBD has lost most peaks of antibody binding relative to the original RBD.
-Exploration of the mutations using the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} indicates that mutations at sites 484, 446, and 417 are the biggest drivers of this antigenic change, although other mutations also contribute.
+Exploration of the mutations using the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} (last accessed March-11-2022) indicates that mutations at sites 484, 446, and 417 are the biggest drivers of this antigenic change, although other mutations also contribute.
 The residual peaks in the map suggests the remaining antibody activity against the Omicron variant RBD could be further eroded by mutations at sites like 346, 378, 444, and 504.
@@
 We encourage interactive use of the estimator so it acts as a visual aid to augment human interpretation.
-However, we note that in addition to the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/}, we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}.
+However, we note that in addition to the interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} (last accessed March-11-2022), we also provide a Python module for batch processing of large numbers of sequences at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/} (last accessed March-11-2022).
 
@@
 \subsection{Code and data availability}
-The most up-to-date code and data used to implement the escape estimator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps}, and the version described in this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/tree/bioRxiv_v1}.
+The most up-to-date code and data used to implement the escape estimator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps} (last accessed March-11-2022), and the version described in this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/tree/bioRxiv_v1} (last accessed March-11-2022).
 
-The data used by the escape estimator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv}, and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv}.
+The data used by the escape estimator are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/main/processed_data/escape_calculator_data.csv} (last accessed March-11-2022), and the version used for this paper are at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_calculator_data.csv} (last accessed March-11-2022).
 
-A downloadable HTML version of the escape estimator is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/raw/main/docs/_includes/escape_calc_chart.html}, with the version described in this paper at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/docs/_includes/escape_calc_chart.html}.
+A downloadable HTML version of the escape estimator is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/raw/main/docs/_includes/escape_calc_chart.html} (last accessed March-11-2022), with the version described in this paper at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/docs/_includes/escape_calc_chart.html} (last accessed March-11-2022).
 
 \subsection{Interactive versions of figures}
-Interactive versions of all of the figures in this paper are at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/}.
+Interactive versions of all of the figures in this paper are at \url{https://jbloomlab.github.io/RBD_escape_calculator_paper/} (last accessed March-11-2022).
 These figures allow mousing over points to see details, etc.
@@
 \subsection{Deep mutational scanning data used by the estimator}
-The experimental data used by the escape estimator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955}.
+The experimental data used by the escape estimator are drawn from seven previously published deep mutational scanning studies~\citep{greaney2021complete,greaney2021mapping,starr2021prospective,starr2021complete,starr2021sars,dong2021genetic,tortorici2021broad} and one unpublished dataset available at \url{https://github.com/jbloomlab/SARS-CoV-2-RBD_MAP_COV2-2955} (last accessed March-11-2022).
 In total, these studies contain data for 36 monoclonal antibodies.
@@
 The majority of the antibodies were originally isolated from cohorts of individuals infected with SARS-CoV-2 in the first half of 2020, and were initially characterized by the Crowe lab~\citep{zost2020potently}, Nussenzweig lab~\citep{robbiani2020convergent}, or Vir Biotechnology~\citep{piccoli2020mapping}, with a few additional antibodies coming from commercial synthesis based on previously reported sequences~\citep{hansen2020studies,jones2021neutralizing,shi2020human}.
-The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper).
+The full deep mutational scanning data for all these antibodies are interactively displayed at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/} (last accessed March-11-2022) and available in raw form at \url{https://raw.githubusercontent.com/jbloomlab/SARS2_RBD_Ab_escape_maps/main/processed_data/escape_data.csv} (see \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/blob/bioRxiv_v1/processed_data/escape_data.csv} for a stable version of the raw data corresponding to that used in this paper) (last accessed March-11-2022).
 
@@
 
-We normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
+We normalize the site-level escape metrics for each antibody to account for different strengths of antibody selection in different experiments using the approach described in \citet{greaney2021comprehensive} and implemented in \url{https://jbloomlab.github.io/dmslogo/dmslogo.utils.html#dmslogo.utils.AxLimSetter} (last accessed March-11-2022) with \texttt{min\_upperlim=1} and \texttt{max\_from\_quantile=(0.5, 0.05)}: essentially this corresponds to scaling the site escape values for each antibody so that a value of one corresponds to the larger of the maximum escape at a site or 20 times the median value across sites.
 Specifically, the normalization is done so that for each antibody $a$, the maximum escape $x_{a,r}$ at any site $r$ is set so that $1 = \max\left(\max_r \left[x_{a,r}\right], 20\tilde{x}_{a,r} \right)$ where $\tilde{x}_{a,r}$ is the median of $x_{a,r}$ across sites $r$ for antibody $a$.
@@
 \subsection{Implementation of interactive estimator}
-The interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} is imlemented using Altair (\url{https://altair-viz.github.io/})~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.
+The interactive estimator at \url{https://jbloomlab.github.io/SARS2_RBD_Ab_escape_maps/escape-calc/} (last accessed March-11-2022) is imlemented using Altair~\citep{VanderPlas2018}, which is in turn built upon Vega-Lite~\citep{Satyanarayan2017}.
 
 \subsection{Python module with batch-mode estimator}
-A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/}, and has all the same options as the interactive estimator.
+A Python module that implements the calculations is at \url{https://github.com/jbloomlab/SARS2_RBD_Ab_escape_maps/} (last accessed March-11-2022), and has all the same options as the interactive estimator.
 
@@
 For each study cohort, we computed the geometric mean fold change in neutralization titer over all subjects.
-The numerical compiled data are at \url{https://github.com/jbloomlab/RBD_escape_calculator_paper/tree/main/results/neut_studies}.
+The numerical compiled data are at \url{https://github.com/jbloomlab/RBD_escape_calculator_paper/tree/main/results/neut_studies} (last accessed March-11-2022).
 

### 2022-03-11 paper/paper_with_refs.tex
made version with refs in LaTex

--- 2022-03-11 paper/paper.tex
@@
 
-\bibliography{references}
+\begin{thebibliography}{30}
+
+\bibitem[{Barnes {\em et~al.\/}}(2020){Barnes, Jette, Abernathy, Dam, Esswein,
+  Gristick, Malyutin, Sharaf, Huey-Tubman, Lee, {\em
+  et~al.\/}}]{barnes2020sars}
+{Barnes, C.~O., C.~A. Jette, M.~E. Abernathy, K.-M.~A. Dam, S.~R. Esswein, {\em
+  et~al.\/}}, 2020 {SARS-CoV-2} neutralizing antibody structures inform
+  therapeutic strategies. Nature {\bf 588}: 682--687.
+
+\bibitem[{Chen {\em et~al.\/}}(2021){Chen, Gilchuk, Zost, Suryadevara, Winkler,
+  Cabel, Binshtein, Sutton, Rodriguez, Day, {\em
+  et~al.\/}}]{chen2021convergent}
+{Chen, E.~C., P.~Gilchuk, S.~J. Zost, N.~Suryadevara, E.~S. Winkler, {\em
+  et~al.\/}}, 2021 Convergent antibody responses to the {SARS-CoV-2} spike
+  protein in convalescent and vaccinated individuals. Cell Reports {\bf 36}:
+  109604.
+
+\bibitem[{Cobey {\rm and} Hensley}(2017)]{cobey2017immune}
+{Cobey, S. {\rm and} S.~E. Hensley}, 2017 Immune history and influenza virus
+  susceptibility. Current opinion in virology {\bf 22}: 105--111.
+
+\bibitem[{de~Oliveira}(2021)]{deoliveira2021tweet}
+{de~Oliveira, T.}, 2021
+  \url{https://twitter.com/Tuliodna/status/1463911554538160130}.
+
+\bibitem[{Dong {\em et~al.\/}}(2021){Dong, Zost, Greaney, Starr, Dingens, Chen,
+  Chen, Case, Sutton, Gilchuk, {\em et~al.\/}}]{dong2021genetic}
+{Dong, J., S.~J. Zost, A.~J. Greaney, T.~N. Starr, A.~S. Dingens, {\em
+  et~al.\/}}, 2021 Genetic and structural basis for {SARS-CoV-2} variant
+  neutralization by a two-antibody cocktail. Nature Microbiology {\bf 6}:
+  1233--1244.
+
+\bibitem[{Eguia {\em et~al.\/}}(2021){Eguia, Crawford, Stevens-Ayers,
+  Kelnhofer-Millevolte, Greninger, Englund, Boeckh, {\rm and}
+  Bloom}]{eguia2021human}
+{Eguia, R.~T., K.~H. Crawford, T.~Stevens-Ayers, L.~Kelnhofer-Millevolte, A.~L.
+  Greninger, {\em et~al.\/}}, 2021 A human coronavirus evolves antigenically to
+  escape antibody immunity. PLoS Pathogens {\bf 17}: e1009453.
+
+\bibitem[{Einav {\rm and} Bloom}(2020)]{einav2020two}
+{Einav, T. {\rm and} J.~D. Bloom}, 2020 When two are better than one: Modeling
+  the mechanisms of antibody mixtures. PLoS Computational Biology {\bf 16}:
+  e1007830.
+
+\bibitem[{Greaney {\em et~al.\/}}(2021{a}){Greaney, Loes, Crawford, Starr,
+  Malone, Chu, {\rm and} Bloom}]{greaney2021comprehensive}
+{Greaney, A.~J., A.~N. Loes, K.~H. Crawford, T.~N. Starr, K.~D. Malone, {\em
+  et~al.\/}}, 2021{a} Comprehensive mapping of mutations in the {SARS-CoV-2}
+  receptor-binding domain that affect recognition by polyclonal human plasma
+  antibodies. Cell Host \& Microbe {\bf 29}: 463--476.
+
+\bibitem[{Greaney {\em et~al.\/}}(2021{b}){Greaney, Starr, Barnes, Weisblum,
+  Schmidt, Caskey, Gaebler, Cho, Agudelo, Finkin, {\em
+  et~al.\/}}]{greaney2021mapping}
+{Greaney, A.~J., T.~N. Starr, C.~O. Barnes, Y.~Weisblum, F.~Schmidt, {\em
+  et~al.\/}}, 2021{b} Mapping mutations to the {SARS-CoV-2 RBD} that escape
+  binding by different classes of antibodies. Nature communications {\bf 12}:
+  1--14.
+
+\bibitem[{Greaney {\em et~al.\/}}(2022){Greaney, Starr, Eguia, Loes, Khan,
+  Karim, Cele, Bowen, Logue, Corti, {\em et~al.\/}}]{greaney2021sars}
+{Greaney, A.~J., T.~N. Starr, R.~T. Eguia, A.~N. Loes, K.~Khan, {\em
+  et~al.\/}}, 2022 A {SARS-CoV-2} variant elicits an antibody response with a
+  shifted immunodominance hierarchy. PLoS Pathogens {\bf 18:e1010248}.
+
+\bibitem[{Greaney {\em et~al.\/}}(2021{c}){Greaney, Starr, Gilchuk, Zost,
+  Binshtein, Loes, Hilton, Huddleston, Eguia, Crawford, {\em
+  et~al.\/}}]{greaney2021complete}
+{Greaney, A.~J., T.~N. Starr, P.~Gilchuk, S.~J. Zost, E.~Binshtein, {\em
+  et~al.\/}}, 2021{c} Complete mapping of mutations to the {SARS-CoV-2} spike
+  receptor-binding domain that escape antibody recognition. Cell Host \&
+  Microbe {\bf 29}: 44--57.
+
+\bibitem[{Hansen {\em et~al.\/}}(2020){Hansen, Baum, Pascal, Russo, Giordano,
+  Wloga, Fulton, Yan, Koon, Patel, {\em et~al.\/}}]{hansen2020studies}
+{Hansen, J., A.~Baum, K.~E. Pascal, V.~Russo, S.~Giordano, {\em et~al.\/}},
+  2020 Studies in humanized mice and convalescent humans yield a {SARS-CoV-2}
+  antibody cocktail. Science {\bf 369}: 1010--1014.
+
+\bibitem[{Jones {\em et~al.\/}}(2021){Jones, Brown-Augsburger, Corbett,
+  Westendorf, Davies, Cujec, Wiethoff, Blackbourne, Heinz, Foster, {\em
+  et~al.\/}}]{jones2021neutralizing}
+{Jones, B.~E., P.~L. Brown-Augsburger, K.~S. Corbett, K.~Westendorf, J.~Davies,
+  {\em et~al.\/}}, 2021 The neutralizing antibody, {LY-CoV555}, protects
+  against {SARS-CoV-2} infection in nonhuman primates. Science Translational
+  Medicine {\bf 13}.
+
+\bibitem[{Kistler {\rm and} Bedford}(2021)]{kistler2021evidence}
+{Kistler, K.~E. {\rm and} T.~Bedford}, 2021 Evidence for adaptive evolution in
+  the receptor-binding domain of seasonal coronaviruses {OC43} and {229E}.
+  eLife {\bf 10}: e64509.
+
+\bibitem[{Lucas {\em et~al.\/}}(2021){Lucas, Vogels, Yildirim, Rothman, Lu,
+  Monteiro, Gehlhausen, Campbell, Silva, Tabachnikova, {\em
+  et~al.\/}}]{lucas2021impact}
+{Lucas, C., C.~B. Vogels, I.~Yildirim, J.~E. Rothman, P.~Lu, {\em et~al.\/}},
+  2021 Impact of circulating {SARS-CoV-2} variants on {mRNA} vaccine-induced
+  immunity. Nature pp. 1--7.
+
+\bibitem[{{NGS-SA}}(2021)]{ngs-sa2021update}
+{{NGS-SA}}, 2021 {SARS-CoV-2 sequencing update 26 November 2021 by the Network
+  for Genomic Surveillance in South Africa}.
+  \url{https://www.nicd.ac.za/wp-content/uploads/2021/11/Update-of-SA-sequencing-data-from-GISAID-26-Nov_Final.pdf}.
+
+\bibitem[{Piccoli {\em et~al.\/}}(2020){Piccoli, Park, Tortorici,
+  Czudnochowski, Walls, Beltramello, Silacci-Fregni, Pinto, Rosen, Bowen, {\em
+  et~al.\/}}]{piccoli2020mapping}
+{Piccoli, L., Y.-J. Park, M.~A. Tortorici, N.~Czudnochowski, A.~C. Walls, {\em
+  et~al.\/}}, 2020 Mapping neutralizing and immunodominant sites on the
+  {SARS-CoV-2} spike receptor-binding domain by structure-guided
+  high-resolution serology. Cell {\bf 183}: 1024--1042.
+
+\bibitem[{Robbiani {\em et~al.\/}}(2020){Robbiani, Gaebler, Muecksch, Lorenzi,
+  Wang, Cho, Agudelo, Barnes, Gazumyan, Finkin, {\em
+  et~al.\/}}]{robbiani2020convergent}
+{Robbiani, D.~F., C.~Gaebler, F.~Muecksch, J.~C. Lorenzi, Z.~Wang, {\em
+  et~al.\/}}, 2020 Convergent antibody responses to {SARS-CoV-2} in
+  convalescent individuals. Nature {\bf 584}: 437--442.
+
+\bibitem[{Satyanarayan {\em et~al.\/}}(2017){Satyanarayan, Moritz,
+  Wongsuphasawat, {\rm and} Heer}]{Satyanarayan2017}
+{Satyanarayan, A., D.~Moritz, K.~Wongsuphasawat, {\rm and} J.~Heer}, 2017
+  {Vega-Lite}: A grammar of interactive graphics. IEEE transactions on
+  visualization and computer graphics {\bf 23}: 341--350.
+
+\bibitem[{Schmidt {\em et~al.\/}}(2021){Schmidt, Weisblum, Rutkowska, Poston,
+  DaSilva, Zhang, Bednarski, Cho, Schaefer-Babajew, Gaebler, {\em
+  et~al.\/}}]{schmidt2021high}
+{Schmidt, F., Y.~Weisblum, M.~Rutkowska, D.~Poston, J.~DaSilva, {\em
+  et~al.\/}}, 2021 High genetic barrier to {SARS-CoV-2} polyclonal neutralizing
+  antibody escape. Nature pp. 1--5.
+
+\bibitem[{Shi {\em et~al.\/}}(2020){Shi, Shan, Duan, Chen, Liu, Song, Song, Bi,
+  Han, Wu, {\em et~al.\/}}]{shi2020human}
+{Shi, R., C.~Shan, X.~Duan, Z.~Chen, P.~Liu, {\em et~al.\/}}, 2020 A human
+  neutralizing antibody targets the receptor-binding site of sars-cov-2. Nature
+  {\bf 584}: 120--124.
+
+\bibitem[{Starr {\em et~al.\/}}(2021{a}){Starr, Czudnochowski, Liu, Zatta,
+  Park, Addetia, Pinto, Beltramello, Hernandez, Greaney, {\em
+  et~al.\/}}]{starr2021sars}
+{Starr, T.~N., N.~Czudnochowski, Z.~Liu, F.~Zatta, Y.-J. Park, {\em et~al.\/}},
+  2021{a} {SARS-CoV-2 RBD} antibodies that maximize breadth and resistance to
+  escape. Nature {\bf 597}: 97--102.
+
+\bibitem[{Starr {\em et~al.\/}}(2021{b}){Starr, Greaney, Addetia, Hannon,
+  Choudhary, Dingens, Li, {\rm and} Bloom}]{starr2021prospective}
+{Starr, T.~N., A.~J. Greaney, A.~Addetia, W.~W. Hannon, M.~C. Choudhary, {\em
+  et~al.\/}}, 2021{b} Prospective mapping of viral mutations that escape
+  antibodies used to treat {COVID-19}. Science {\bf 371}: 850--854.
+
+\bibitem[{Starr {\em et~al.\/}}(2021{c}){Starr, Greaney, Dingens, {\rm and}
+  Bloom}]{starr2021complete}
+{Starr, T.~N., A.~J. Greaney, A.~S. Dingens, {\rm and} J.~D. Bloom}, 2021{c}
+  {Complete map of SARS-CoV-2 RBD mutations that escape the monoclonal antibody
+  LY-CoV555 and its cocktail with LY-CoV016}. Cell Reports Medicine {\bf 2}:
+  100255.
+
+\bibitem[{Tortorici {\em et~al.\/}}(2021){Tortorici, Czudnochowski, Starr,
+  Marzi, Walls, Zatta, Bowen, Jaconi, Di~Iulio, Wang, {\em
+  et~al.\/}}]{tortorici2021broad}
+{Tortorici, M.~A., N.~Czudnochowski, T.~N. Starr, R.~Marzi, A.~C. Walls, {\em
+  et~al.\/}}, 2021 Broad sarbecovirus neutralization by a human monoclonal
+  antibody. Nature {\bf 597}: 103--108.
+
+\bibitem[{Uriu {\em et~al.\/}}(2021){Uriu, Kimura, Shirakawa, Takaori-Kondo,
+  Nakada, Kaneda, Nakagawa, {\rm and} Sato}]{uriu2021neutralization}
+{Uriu, K., I.~Kimura, K.~Shirakawa, A.~Takaori-Kondo, T.-a. Nakada, {\em
+  et~al.\/}}, 2021 Neutralization of the {SARS-CoV-2 Mu} variant by
+  convalescent and vaccine serum. New England Journal of Medicine .
+
+\bibitem[{VanderPlas {\em et~al.\/}}(2018){VanderPlas, Granger, Heer, Moritz,
+  Wongsuphasawat, Satyanarayan, Lees, Timofeev, Welsh, {\rm and}
+  Sievert}]{VanderPlas2018}
+{VanderPlas, J., B.~Granger, J.~Heer, D.~Moritz, K.~Wongsuphasawat, {\em
+  et~al.\/}}, 2018 Altair: Interactive statistical visualizations for {Python}.
+  Journal of Open Source Software {\bf 3}: 1057.
+
+\bibitem[{Wang {\em et~al.\/}}(2021){Wang, Nair, Liu, Iketani, Luo, Guo, Wang,
+  Yu, Zhang, Kwong, {\em et~al.\/}}]{wang2021antibody}
+{Wang, P., M.~S. Nair, L.~Liu, S.~Iketani, Y.~Luo, {\em et~al.\/}}, 2021
+  Antibody resistance of {SARS-CoV-2} variants {B.1.351} and {B.1.1.7}. Nature
+  {\bf 593}: 130--135.
+
+\bibitem[{Yuan {\em et~al.\/}}(2020){Yuan, Liu, Wu, Lee, Zhu, Zhao, Huang, Yu,
+  Hua, Tien, {\em et~al.\/}}]{yuan2020structural}
+{Yuan, M., H.~Liu, N.~C. Wu, C.-C.~D. Lee, X.~Zhu, {\em et~al.\/}}, 2020
+  Structural basis of a shared antibody response to {SARS-CoV-2}. Science {\bf
+  369}: 1119--1123.
+
+\bibitem[{Zost {\em et~al.\/}}(2020){Zost, Gilchuk, Case, Binshtein, Chen,
+  Nkolola, Sch{\"a}fer, Reidy, Trivette, Nargi, {\em
+  et~al.\/}}]{zost2020potently}
+{Zost, S.~J., P.~Gilchuk, J.~B. Case, E.~Binshtein, R.~E. Chen, {\em
+  et~al.\/}}, 2020 Potently neutralizing and protective human antibodies
+  against {SARS-CoV-2}. Nature {\bf 584}: 443--449.
+
+\end{thebibliography}
+
 }
